---
title: "MOTIVATE mapping - Calculating habitat extent from Mahalanobis rasters"
author: "Alicia Valdés"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: 4
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  word_document:
    toc: yes
    toc_depth: '4'
editor_options:
  chunk_output_type: inline
---

This notebook calculates habitat extent from Mahalanobis rasters.

# Load libraries

```{r}
library(terra)
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(sf)
library(readr)
library(here)
library(exactextractr)
library(furrr)
library(tmap)
library(ggplot2)
library(RColorBrewer)
library(colorspace)
```

# Load data

Load points from ReSurvey Europe observations from validated plots. Points are in WGS84 (EPSG:4326) - convert to ETRS89-extended / LAEA Europe (EPSG:3035) for better area calculations.

```{r}
points <- read_csv(here("..", "MOTIVATE_validation", "data", "clean", "initial_dataset_validation.csv")) %>%
  # Keep only plot observations from validated plots
  filter(valid_plot == TRUE) %>%
  # Remove unneeded columns
  dplyr::select(- valid_obs, - valid_plot) %>%
  st_as_sf(coords = c("Lon_updated", "Lat_updated"), crs = 4326) %>%
  st_transform(3035)
st_crs(points)$Name
```

Create a 500-m buffer around each point.

```{r}
buffers <- points %>% st_buffer(dist = 500, nQuadSegs = 100) # nQuadSegs to make smoother circles
st_crs(buffers)$Name
```

Check that buffer areas are ok (should be around 785398 m² for a 500-m radius circle).

```{r}
buffers$buffer_area <- as.numeric(sf::st_area(buffers))
```

Load raster files with Mahalanobis distance. 

```{r}
raster_files <- list.files(
  "C:/Data/MOTIVATE/MOTIVATE_RS_data/mapping/mahalanobis_260508_max_1pixel_expBuffer",
  pattern = "\\.tif$", full.names = TRUE, recursive = TRUE)
```

```{r}
raster_files_additional <- list.files(
  "C:/Data/MOTIVATE/MOTIVATE_RS_data/mapping/mahalanobis_260713_max_1pixel_expBuffer_additional",
  pattern = "\\.tif$", full.names = TRUE, recursive = TRUE)
```

# Reproject rasters

Rasters are in WGS84 (EPSG:4326) - convert to ETRS89-extended / LAEA Europe (EPSG:3035) for better area calculations.

```{r message=FALSE}
# ---- OUTPUT FOLDERS ----
output_dir <- "C:/Data/MOTIVATE/MOTIVATE_RS_data/mapping/mahalanobis_reproj"

# Create output folder if needed
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

# ---- LOOP OVER RASTERS ----
for (f in raster_files) {
  
  # Build output filename FIRST
  out_name <- file.path(
    output_dir,
    paste0(tools::file_path_sans_ext(basename(f)), "_LAEA.tif")
  )
  
  # ✅ Skip if already exists
  if (file.exists(out_name)) {
    message("Skipping (already exists): ", basename(out_name))
    next
  }
  
  # Otherwise process
  r <- rast(f)
  
  r_laea <- project(r, "EPSG:3035")
  
  writeRaster(r_laea, out_name, overwrite = FALSE)
}
```

```{r message=FALSE}
# ---- OUTPUT FOLDERS ----
output_dir <- "C:/Data/MOTIVATE/MOTIVATE_RS_data/mapping/mahalanobis_reproj"

# Create output folder if needed
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

# ---- LOOP OVER RASTERS ----
for (f in raster_files_additional) {
  
  # Build output filename FIRST
  out_name <- file.path(
    output_dir,
    paste0(tools::file_path_sans_ext(basename(f)), "_LAEA.tif")
  )
  
  # ✅ Skip if already exists
  if (file.exists(out_name)) {
    message("Skipping (already exists): ", basename(out_name))
    next
  }
  
  # Otherwise process
  r <- rast(f)
  
  r_laea <- project(r, "EPSG:3035")
  
  writeRaster(r_laea, out_name, overwrite = FALSE)
}
```

# Load reprojected rasters

First remove old rasters in WGS84.

```{r}
rm(raster_files)
rm(raster_files_additional)
```

Now load new reprojected rasters in ETRS89.

```{r}
raster_files_proj <- list.files(
  "C:/Data/MOTIVATE/MOTIVATE_RS_data/mapping/mahalanobis_reproj", 
  pattern = "\\.tif$", full.names = TRUE, recursive = TRUE)
```

```{r}
crs(rast(raster_files_proj[[1]]), describe = TRUE)
```

# Get habitat exents from each raster

## Extract PlotObservationID from raster filenames

```{r}
get_id <- function(x) {
  str_extract(x, "(?<=Obs)\\d+") %>% as.numeric()
}
```

## Define percentiles to use

```{r}
percentiles <- c(0.5, 0.7, 0.9)
```

## Function to process each raster

```{r}
process_raster_fractional <- function(f) {
  
  # -------------------------------------------------------------
  # 1. IDENTIFY PLOT AND LOAD RASTER
  # -------------------------------------------------------------
  # Extract PlotObservationID from filename
  id <- get_id(f)
  
  # Load raster (Mahalanobis distance surface for one plot)
  r <- rast(f)
  
  
  # -------------------------------------------------------------
  # 2. SELECT CORRESPONDING BUFFER
  # -------------------------------------------------------------
  # Each raster corresponds to one PlotObservationID → subset matching buffer
  buf <- buffers %>%
    dplyr::filter(PlotObservationID == id)
  
  # If no matching buffer found, exit function
  if (nrow(buf) == 0) return(NULL)
  
  # Ensure CRS compatibility between raster and buffer
  buf <- st_transform(buf, crs(r))
  st_crs(buf) <- st_crs(r)
  
  
  # -------------------------------------------------------------
  # 3. FRACTIONAL EXTRACTION (FOR STATISTICS)
  # -------------------------------------------------------------
  # Convert terra raster → raster object for exactextractr
  r_raster <- raster::raster(r)
  
  # Extract pixel values within buffer using fractional coverage
  # Returns:
  #  - value: pixel Mahalanobis distance
  #  - coverage_fraction: proportion of pixel inside buffer
  df <- suppressWarnings(
    exact_extract(r_raster, buf)[[1]]
  ) %>%
    dplyr::filter(!is.na(value))
  
  if (nrow(df) == 0) return(NULL)
  
  
  # -------------------------------------------------------------
  # 4. COMPUTE BASIC METRICS
  # -------------------------------------------------------------
  
  # Area of a full raster pixel (m²)
  pixel_area <- prod(res(r))
  
  # Fractional number of pixels in buffer
  total_weight <- sum(df$coverage_fraction)
  
  # True buffer area (based on raster coverage, not vector geometry)
  area_buffer <- total_weight * pixel_area
  
  
  # -------------------------------------------------------------
  # 5. COMPUTE THRESHOLDS
  # -------------------------------------------------------------
  # Thresholds are percentiles of Mahalanobis distance distribution
  thr <- quantile(df$value, probs = percentiles, na.rm = TRUE)
  
  
  # -------------------------------------------------------------
  # 6. PREPARE RASTER (OPTIMIZED)
  # -------------------------------------------------------------
  # ✅ Crop and mask raster to buffer → reduces computation massively
  r_crop <- crop(r, vect(buf))
  buf_vect <- terra::vect(buf)[, 0]  # ✅ drop ALL attributesr
  r_mask <- mask(r_crop, buf_vect)
  
  
  # -------------------------------------------------------------
  # 7. LOOP OVER THRESHOLDS (PERCENTILES)
  # -------------------------------------------------------------
  results <- purrr::map_dfr(seq_along(thr), function(i) {
    
    threshold <- thr[i]          # Mahalanobis threshold value
    perc_val <- percentiles[i]  # percentile (e.g. 0.5, 0.7, 0.9)
    
    
    # -----------------------------------------------------------
    # 7.1 CLASSIFY RASTER INTO HABITAT / NON-HABITAT (FAST)
    # -----------------------------------------------------------
    # NOTE:
    # value ≥ threshold = more similar → habitat
    # value < threshold = less similar → non-habitat
    # ✅ classification done in raster space (much faster)
    r_class <- ifel(r_mask >= threshold, 1L, 0L)
    
    
    # -----------------------------------------------------------
    # 7.2 CONVERT CLASSIFIED RASTER TO POLYGONS
    # -----------------------------------------------------------
    # ✅ Only final classes are polygonized → very few polygons
    pol_sf_thr <- as.polygons(r_class, dissolve = TRUE) %>%
      st_as_sf() %>%
      dplyr::rename(extent = 1) %>%   # rename raster value column
      dplyr::select(extent, geometry) # ✅ KEEP ONLY THESE

    
    # -----------------------------------------------------------
    # 7.3 CLIP POLYGONS TO BUFFER (IMPORTANT)
    # -----------------------------------------------------------
    # ✅ Removes pixel parts outside buffer
    # ✅ Restores true circular boundary
    pol_sf_thr <- suppressWarnings(
      st_intersection(pol_sf_thr, buf)
      )
    
    # ✅ VERY IMPORTANT: remove buffer attributes introduced by intersection
    pol_sf_thr <- pol_sf_thr %>%
      dplyr::select(extent, geometry)
    
    pol_sf_thr <- st_make_valid(pol_sf_thr)
    
    
    # -----------------------------------------------------------
    # 7.4 COMPUTE AREAS (FRACTIONAL, FROM EXACTEXTRACTR)
    # -----------------------------------------------------------
    # IMPORTANT:
    # Use exactextractr-based areas for correct fractional accounting
    
    hab_weight <- sum(df$coverage_fraction[df$value >= threshold], na.rm = TRUE)
    non_weight <- sum(df$coverage_fraction[df$value < threshold], na.rm = TRUE)
    
    area_hab <- hab_weight * pixel_area
    area_non <- non_weight * pixel_area
    
    
    # -----------------------------------------------------------
    # 7.5 RETURN RESULTS
    # -----------------------------------------------------------
    tibble(
      PlotObservationID = id,
      percentile = perc_val,
      threshold_dist = threshold,
      n_pix_buffer = total_weight,
      n_pix_extent = hab_weight,
      n_pix_out = non_weight,
      buffer_m2 = area_buffer,
      extent_m2 = area_hab,
      out_m2 = area_non,
      extent_perc = area_hab / area_buffer * 100
    ) %>%
      dplyr::mutate(geometry = list(pol_sf_thr))
  })
  
  
  # -------------------------------------------------------------
  # 8. RETURN OUTPUT
  # -------------------------------------------------------------
  return(results)
}
```

```{r}
# load existing results
results_all <- readRDS(here("output", "extent_results_fractional.rds"))

# IDs already processed
done_ids <- unique(results_all$PlotObservationID)

# IDs from rasters
raster_ids <- sapply(raster_files_proj, get_id)

# select rasters not yet processed
rasters_to_run <- raster_files_proj[!raster_ids %in% done_ids]

# Print what will run
message("New rasters to process: ", length(rasters_to_run))
message("Already processed: ", length(done_ids))

# Safety check
if (length(rasters_to_run) == 0) {
  message("Nothing to process!")
} else {
  
  n_total <- length(rasters_to_run)
  
  new_results <- purrr::map_dfr(
    seq_along(rasters_to_run),
    function(i) {
      
      f <- rasters_to_run[i]
      id <- get_id(f)
      
      message(sprintf("Processing %d/%d - Obs: %s", i, n_total, id))
      
      tryCatch(
        process_raster_fractional(f),
        error = function(e) {
          message("ERROR in ID ", id, ": ", e$message)
          return(NULL)
        }
      )
    }
  )
  
  # combine
  results_all <- dplyr::bind_rows(results_all, new_results)
  
  # save updated results
  saveRDS(results_all, here("output", "extent_results_fractional.rds"))
}
```

## Get results table

```{r}
results_tbl <- results_all %>%
  dplyr::select(-geometry)
```

```{r}
results_tbl
```

# Keep only observations from validated plots

```{r}
extent_val_plots <- points %>% 
  left_join(results_tbl)
```

```{r}
extent_val_plots <- extent_val_plots %>% arrange(plotID, percentile, year)
```

## Number of time series

```{r}
extent_val_plots %>% distinct(plotID) %>% nrow()
```

## Save csv

```{r}
extent_val_plots_csv <- extent_val_plots %>%
  mutate(
    x = st_coordinates(.)[,1],
    y = st_coordinates(.)[,2]
  ) %>%
  st_drop_geometry()

write_csv(extent_val_plots_csv, here("data", "clean", "extent_val_plots.csv"))
```

# Export to shapefiles

## Get geomtries

```{r}
# geometries (unnest)
all_polygons <- results_all %>%
  tidyr::unnest(geometry) %>%
  sf::st_as_sf() %>%
  dplyr::select(
    PlotObservationID,
    percentile,
    threshold_dist,
    n_pix_buffer,
    n_pix_extent,
    n_pix_out,
    buffer_m2,
    extent_m2,
    out_m2,
    extent_perc,
    extent,
    geometry
    )
```

## Define directory to save shapefiles

```{r}
out_shapes <- here("output", "shapefiles")
dir.create(out_shapes, recursive = TRUE, showWarnings = FALSE)
```

## Export

```{r}
walk(unique(all_polygons$percentile), function(p) {
  
  shp <- all_polygons %>%
    dplyr::filter(percentile == p)
  
  # Adapt names for shapefile
  names(shp) <- make.names(names(shp), unique = TRUE)
  names(shp) <- substr(names(shp), 1, 10)
  
  out_name <- file.path(
    out_shapes,
    paste0("extent_p", p * 100, ".shp")
  )
  
  st_write(
    shp,
    out_name,
    delete_layer = TRUE,
    quiet = TRUE
  )
})
```

# Change in habitat extent

## Keep only first and last year for each plot

```{r}
extent_val_plots_first_last <- extent_val_plots %>%
  filter(year == first_year | year == last_year) %>%
  distinct()
```

## Remove time series with only one year

```{r}
extent_val_plots_first_last <- extent_val_plots_first_last %>% 
  filter(n_resurveys > 1)
```

## Remove time series with missing extent data

```{r}
extent_val_plots_first_last <- extent_val_plots_first_last %>%
  group_by(plotID) %>%
  filter(!any(is.na(extent_perc))) %>%
  ungroup()
```

## Assign EUNIS habitat type to each time series

```{r}
extent_val_plots_first_last <- extent_val_plots_first_last %>%
  rename(EUNIS_obs = EUNIS1) %>%
  group_by(plotID) %>%
  mutate(
    EUNIS_plot = if_else(
      n_distinct(EUNIS_obs) > 1,
      "Change",
      first(EUNIS_obs)
    )
  ) %>%
  ungroup()
```

## Number of time series

```{r}
extent_val_plots_first_last %>% distinct(plotID) %>% nrow()
```

## Calculate change in extent

```{r eval=FALSE, include=FALSE}
extent_change <- extent_val_plots_first_last %>%
  group_by(plotID, percentile) %>%
  summarise(
    EUNIS_plot = first(EUNIS_plot),
    first_year = first(first_year),
    last_year = first(last_year),
    extent_first = extent_perc[year == first_year][1],
    extent_last  = extent_perc[year == last_year][1],
    absolute_change = extent_last - extent_first,
    relative_change = 100 * (extent_last - extent_first) / extent_first,
    annual_change = (extent_last - extent_first) / (last_year - first_year),
    .groups = "drop"
    ) %>%
  st_drop_geometry()
```

## Save csv

```{r eval=FALSE, include=FALSE}
write.csv(
  st_drop_geometry(extent_change),
  here("data", "clean", "extent_change.csv"),
  row.names = FALSE
)
```

# Assign a biogeographical region to each plot

Extract the latest coordinate pair per plot (sometimes coordinates are different among different observations of the same plot):

```{r eval=FALSE, include=FALSE}
# One point per plotID, from the last observation year
plot_coords <- extent_val_plots %>%
  filter(year == last_year) %>%
  distinct(plotID, .keep_all = TRUE) %>%
  dplyr::select(plotID, geometry)

# Join to extent_change
extent_change_sf <- extent_change %>%
  left_join(plot_coords, by = "plotID") %>%
  st_as_sf()
```

Read shapefile with info on bioregions:

```{r}
biogeo <- st_read("C:/Data/PAs/Biogeoregions/PAs/N2K_BiogeoUNITS.shp")
```

```{r}
# Ensure same CRS
biogeo <- st_transform(
  biogeo,
  st_crs(extent_change_sf)
)

# Spatial join
extent_change_sf <- extent_change_sf %>%
  st_join(
    biogeo %>%
      dplyr::select(BIOGEO, UNIT),
    join = st_intersects,
    left = TRUE
  )
```

## Map of plots in bioregions

```{r}
# --------------------------------------------------
# One location per plot
# --------------------------------------------------

plot_locations <- extent_change_sf %>%
  distinct(plotID, .keep_all = TRUE) %>%
  mutate(
    BIOUNIT = paste(BIOGEO, UNIT, sep = "_")
  )

# --------------------------------------------------
# Add BIOUNIT to polygons
# --------------------------------------------------

biogeo_plot <- biogeo %>%
  mutate(
    BIOUNIT = paste(BIOGEO, UNIT, sep = "_")
  )

# --------------------------------------------------
# Sampled units
# --------------------------------------------------

sampled_units <- unique(plot_locations$BIOUNIT)

biogeo_plot <- biogeo_plot %>%
  mutate(
    sampled = BIOUNIT %in% sampled_units
  )

biogeo_sampled <- biogeo_plot %>%
  filter(sampled)

# --------------------------------------------------
# Zoom extent
# --------------------------------------------------

bb <- st_bbox(biogeo_sampled)

# --------------------------------------------------
# Base colours by BIOGEO
# --------------------------------------------------

biogeo_cols <- c(
  ALP = "#2166AC",  # blue
  ATL = "#F4A300",  # orange
  CON = "#1B9E77",  # green
  MED = "#D73027",  # red
  PAN = "#7B3294"   # purple
)

# --------------------------------------------------
# Create shades within each BIOGEO
# --------------------------------------------------

biounit_df <- plot_locations %>%
  st_drop_geometry() %>%
  distinct(BIOGEO, UNIT, BIOUNIT) %>%
  arrange(BIOGEO, UNIT)

cols <- c()

for (bg in unique(biounit_df$BIOGEO)) {

  units_bg <- biounit_df %>%
    filter(BIOGEO == bg)

  n_units <- nrow(units_bg)

  pal <- colorRampPalette(
    c(
      lighten(biogeo_cols[bg], amount = 0.35),
      darken(biogeo_cols[bg], amount = 0.15)
    )
  )(n_units)

  cols <- c(
    cols,
    setNames(pal, units_bg$BIOUNIT)
  )
}

# --------------------------------------------------
# Plot
# --------------------------------------------------

ggplot() +

  # Background units
  geom_sf(
    data = biogeo_plot,
    fill = "grey95",
    color = "grey75",
    linewidth = 0.2
  ) +

  # Sampled units
  geom_sf(
    data = biogeo_plot,
    aes(fill = BIOUNIT),
    color = "grey40",
    linewidth = 0.3
  ) +

  # Plot locations
  geom_sf(
    data = plot_locations,
    color = "black",
    size = 0.12,
    alpha = 0.4
  ) +

  scale_fill_manual(
    values = cols,
    name = "Biogeographical unit"
  ) +

  coord_sf(
    xlim = c(bb["xmin"], bb["xmax"]),
    ylim = c(bb["ymin"], bb["ymax"]),
    expand = FALSE
  ) +

  labs(
    title = "Distribution of plots across biogeographical units"
  ) +

  theme_bw() +

  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title = element_blank(),
    legend.position = "right",
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 8),
    plot.title = element_text(
      face = "bold",
      hjust = 0.5
    )
  )
```

## Count of plots in bioregions

```{r}
extent_change_sf %>% st_drop_geometry() %>% dplyr::select(plotID, BIOGEO, UNIT) %>% 
  distinct() %>% count(BIOGEO) %>% arrange(n)
extent_change_sf %>% st_drop_geometry() %>% dplyr::select(plotID, BIOGEO, UNIT) %>% 
  distinct() %>% count(BIOGEO, UNIT) %>% arrange(n)
```

## Save as .rds

```{r eval=FALSE, include=FALSE}
saveRDS(extent_change_sf, here("data", "clean", "extent_change_sf.rds"))
```

## Load .rds

```{r}
extent_change <- readRDS(here("data", "clean", "extent_change_sf.rds"))
```

# Plots

## Change in extent

```{r}
ggplot(extent_change %>% 
         pivot_longer(cols = c(absolute_change:annual_change), 
                      names_to = "change_type"), 
       aes(x = value)) + geom_histogram() +
  facet_grid(rows = vars(change_type), cols = vars(percentile), scales = "free")
```

```{r}
ggplot(extent_change %>% 
         pivot_longer(cols = c(absolute_change:annual_change), 
                      names_to = "change_type") %>%
         # Filter extreme values, look at them later
         filter(value < 25 & value > -25), 
       aes(x = value)) + geom_histogram() +
  facet_grid(rows = vars(change_type), cols = vars(percentile), scales = "free")
```

## Change in extent by habitat type

```{r}
extent_change %>%
  pivot_longer(
    absolute_change:annual_change,
    names_to = "change_type"
  ) %>%
  ggplot(
    aes(
      EUNIS_plot,
      value,
      fill = EUNIS_plot
    )
  ) +
  geom_violin() +
  facet_grid(
    change_type ~ percentile,
    scales = "free_y"
  ) +
  theme_bw() +
  theme(
    legend.position = "none"
  )
```

```{r}
extent_change %>%
  pivot_longer(
    absolute_change:annual_change,
    names_to = "change_type"
  ) %>%
  # Filter extreme values, look at them later
  filter(value < 25 & value > -25) %>%
  ggplot(
    aes(
      EUNIS_plot,
      value,
      fill = EUNIS_plot
    )
  ) +
  geom_violin() +
  facet_grid(
    change_type ~ percentile,
    scales = "free_y"
  ) +
  theme_bw() +
  theme(
    legend.position = "none"
  )
```

## Change in extent by biogeographical region

```{r}
extent_change_sf %>%
  pivot_longer(
    absolute_change:annual_change,
    names_to = "change_type"
  ) %>%
  ggplot(
    aes(
      BIOGEO,
      value,
      fill = BIOGEO
    )
  ) +
  geom_violin() +
  facet_grid(
    change_type ~ percentile,
    scales = "free_y"
  ) +
  theme_bw() +
  theme(
    legend.position = "none"
  )
```

```{r}
extent_change_sf %>%
  pivot_longer(
    absolute_change:annual_change,
    names_to = "change_type"
  ) %>%
  # Filter extreme values, look at them later
  filter(value < 25 & value > -25) %>%
  ggplot(
    aes(
      BIOGEO,
      value,
      fill = BIOGEO
    )
  ) +
  geom_violin() +
  facet_grid(
    change_type ~ percentile,
    scales = "free_y"
  ) +
  theme_bw() +
  theme(
    legend.position = "none"
  )
```

# Maps extreme values of change

## Extract PlotObservationID from raster filenames

```{r}
get_id <- function(x) {
  id <- str_extract(x, "(?<=Obs)\\d+")
  as.numeric(id)
}

raster_ids <- data.frame(
  file = raster_files_proj,
  PlotObservationID = sapply(raster_files_proj, get_id)
)
sum(is.na(raster_ids$PlotObservationID))
```

## Select points

```{r}
plotIDs_extreme <- extent_change %>%
         filter(absolute_change >= 25 | absolute_change <= -25) %>%
  distinct(plotID) %>% pull(plotID)
plots_extreme <- extent_val_plots_first_last %>% 
  filter(plotID %in% plotIDs_extreme)
```

## Export to kml

```{r}
points_kml <- plots_extreme %>%
  st_transform(4326) %>%
  distinct(plotID, .keep_all = TRUE) %>%
  mutate(
    Name = plotID
  )

st_write(
  points_kml["Name"],
  here("kml", "plot_points.kml"),
  delete_dsn = TRUE
)
```

# Maps Mahalanobis

```{r}
tmap_mode("plot")  # for exporting
```
Select one "good" plot:

```{r}
points_map_maha <- points %>% 
  filter(plotID == "South Bohemia floodplain forests - SBoh_forests - 212106")
```

```{r}
output_dir <- "maps_maha"
dir.create(output_dir, showWarnings = FALSE)

plot_ids <- points_map_maha %>%
  distinct(plotID) %>%
  pull(plotID)

for (plot_id in plot_ids) {

  message("Processing: ", plot_id)

  pts_plot <- points_map_maha %>%
    filter(plotID == plot_id)

  years <- sort(unique(pts_plot$year))

  if (length(years) < 2) next

  outfile <- file.path(
    output_dir,
    paste0(
      gsub("[^[:alnum:]]+", "_", plot_id),
      ".png"
    )
  )

  if (file.exists(outfile)) {
    message("Skipping existing map: ", outfile)
    next
  }

  # ======================================================
  # COMMON EXTENT
  # ======================================================

  buf_all <- buffers %>%
    filter(PlotObservationID %in% pts_plot$PlotObservationID) %>%
    st_transform(3857)

  bb_common <- st_bbox(buf_all)

  # ======================================================
  # BASE MAP
  # ======================================================

  obs_base <- pts_plot %>%
    filter(year == years[1]) %>%
    slice(1) %>%
    pull(PlotObservationID)

  pt_base <- points_map_maha %>%
    filter(PlotObservationID == obs_base) %>%
    st_transform(3857)

  buf_base <- buffers %>%
    filter(PlotObservationID == obs_base) %>%
    st_transform(3857)

  title_base <- paste0(
    plot_id,
    " | ",
    min(years),
    "-",
    max(years),
    " | EUNIS: ",
    unique(pts_plot$EUNIS1)[1]
  )

  map_base <-
    tm_shape(buf_base, bbox = bb_common) +
    tm_tiles("Esri.WorldImagery") +
    tm_borders(col = "red", lwd = 2) +
    tm_shape(pt_base) +
    tm_symbols(
      col = "white",
      fill = "white",
      size = 0.5
    ) +
    tm_layout(
      frame = FALSE,
      legend.show = FALSE
    ) +
    tm_title(title_base, size = 0.8)

  # ======================================================
  # MAHALANOBIS MAPS
  # ======================================================

  raster_maps <- list()

  for (yr in years) {

    obs_id <- pts_plot %>%
      filter(year == yr) %>%
      slice(1) %>%
      pull(PlotObservationID)

    pt <- points_map_maha %>%
      filter(PlotObservationID == obs_id) %>%
      st_transform(3857)

    buf <- buffers %>%
      filter(PlotObservationID == obs_id) %>%
      st_transform(3857)

    rfile <- raster_ids$file[
      raster_ids$PlotObservationID == obs_id
    ][1]

    if (is.na(rfile) || !file.exists(rfile))
      next

    r <- rast(rfile)

    r <- project(
      r,
      "EPSG:3857",
      method = "bilinear"
    )

    r <- mask(r, vect(buf))

    raster_maps[[as.character(yr)]] <-

      tm_shape(buf, bbox = bb_common) +
      tm_tiles("Esri.WorldImagery") +

      tm_shape(r) +
      tm_raster(
        col.scale = tm_scale_continuous(values = "-viridis"),
        col_alpha = 0.7,
        col.legend = tm_legend("Similarity")
        ) +
      tm_shape(pt) +
      tm_symbols(
        col = "white",
        fill = "white",
        size = 0.5
      ) +

      tm_layout(
        frame = FALSE,
        legend.show = TRUE
      ) +

      tm_title(
        paste0("Mahalanobis ", yr),
        size = 0.9
      )
  }

  # ======================================================
  # ARRANGE
  # ======================================================
  
  final_map <- tmap_arrange(
    map_base,
    raster_maps[[as.character(years[1])]],
    raster_maps[[as.character(years[2])]],
    ncol = 3,
    nrow = 1
    )
  
  # ======================================================
  # SAVE
  # ======================================================
  
  tmap_save(
    final_map,
    outfile,
    width = 18,
    height = 5,
    units = "in",
    dpi = 300
    )
  
  message("Saved: ", outfile)
  }
```

# Maps extent

```{r}
tmap_mode("plot")  # for exporting
```

Select one "good" plot:

```{r}
points_map_extent <- points %>% 
  filter(plotID == "South Bohemia floodplain forests - SBoh_forests - 212106")
```


```{r}
# output folder
output_dir <- "maps_extent"
dir.create(output_dir, showWarnings = FALSE)

plot_ids <- points_map_extent %>%
  distinct(plotID) %>%
  pull(plotID)

for (plot_id in plot_ids) {

  message("Processing: ", plot_id)

  pts_plot <- points_map_extent %>%
    filter(plotID == plot_id)

  years <- sort(unique(pts_plot$year))

  # skip plots that do not have exactly two years
  if (length(years) != 2) next

  # ---------------------------------------------------------
  # Basemap (use first observation)
  # ---------------------------------------------------------
  obs_base <- pts_plot %>%
    slice(1) %>%
    pull(PlotObservationID)

  pt_base <- pts_plot %>%
    filter(PlotObservationID == obs_base) %>%
    slice(1)

  buf_base <- buffers %>%
    filter(PlotObservationID == obs_base)
  
  eunis <- pt_base$EUNIS1

  map_base <- tm_shape(buf_base) +
    tm_tiles("Esri.WorldImagery") +
    tm_borders(col = "red", lwd = 2) +
    tm_shape(pt_base) +
    tm_symbols(
      col = "white",
      fill = "white",
      size = 0.5
    ) +
    tm_layout(
      frame = FALSE,
      legend.show = FALSE
    ) +
    tm_title(paste0(plot_id, "\nEUNIS: ", eunis, " | Years: ",
                    min(years), "-", max(years)))

  # ---------------------------------------------------------
  # Create percentile maps
  # ---------------------------------------------------------
  extent_maps <- list()

  for (yr in years) {

    obs_id <- pts_plot %>%
      filter(year == yr) %>%
      slice(1) %>%
      pull(PlotObservationID)

    pt <- pts_plot %>%
      filter(PlotObservationID == obs_id) %>%
      slice(1)

    buf <- buffers %>%
      filter(PlotObservationID == obs_id)

    for (p in c(0.5, 0.7, 0.9)) {

      res_id <- results_all %>%
        filter(
          PlotObservationID == obs_id,
          percentile == p
        )

      if (nrow(res_id) == 0) next

      extent_perc <- round(res_id$extent_perc[1], 1)

      pol_sf_hab <- res_id$geometry[[1]] %>%
        filter(extent == 1)

      extent_maps[[paste0(yr, "_", p)]] <-

        tm_shape(buf) +
        tm_tiles("Esri.WorldImagery") +
        tm_borders(col = "red", lwd = 2) +

        tm_shape(pol_sf_hab) +
        tm_fill(
          col = "gold1",
          alpha = 0.6
        ) +

        tm_shape(pt) +
        tm_symbols(
          col = "white",
          fill = "white",
          size = 0.5
        ) +

        tm_layout(
          frame = FALSE,
          legend.show = FALSE
        ) +

        tm_title(
          paste0(
            yr,
            " | P", p * 100,
            "\n",
            round(extent_perc, 1), "%"
          ),
          size = 0.8
        )
    }
  }

  # ---------------------------------------------------------
  # Empty panel under basemap
  # ---------------------------------------------------------
  blank_map <- tm_shape(
    st_as_sf(
      data.frame(x = 0, y = 0),
      coords = c("x", "y"),
      crs = 4326
    )
  ) +
    tm_symbols(size = 0) +
    tm_layout(
      frame = FALSE,
      legend.show = FALSE
    )

  # ---------------------------------------------------------
  # Arrange panels
  # ---------------------------------------------------------
  final_map <- tmap_arrange(
    map_base,
    extent_maps[[paste0(years[1], "_0.5")]],
    extent_maps[[paste0(years[1], "_0.7")]],
    extent_maps[[paste0(years[1], "_0.9")]],

    blank_map,
    extent_maps[[paste0(years[2], "_0.5")]],
    extent_maps[[paste0(years[2], "_0.7")]],
    extent_maps[[paste0(years[2], "_0.9")]],

    ncol = 4,
    nrow = 2
  )

  outfile <- file.path(
    output_dir,
    paste0(
      gsub("[^[:alnum:]]+", "_", plot_id),
      ".png"
    )
  )

  tmap_save(
    final_map,
    outfile,
    width = 14,
    height = 7,
    units = "in",
    dpi = 300
  )

  message("Saved: ", outfile)
}
```

# Session info

```{r}
sessionInfo()
```

