---
title: "MOTIVATE mapping - Processing biodiversity data from Stephan"
author: "Alicia Valdés"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: 4
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  word_document:
    toc: yes
    toc_depth: '4'
editor_options:
  chunk_output_type: inline
---

This notebook processes biodiversity data from Nature Communications paper by Stephan Kambach (https://doi.org/10.1038/s41467-026-72112-5). The data includes time series-specific linear trends in all biodiversity indices, expressed in annual percentage changes and in absolute units, and is openly available at the data repository of the German Centre for Integrated Biodiversity Research (iDiv) Halle-Jena-Leipzig (idata.idiv.de/ddm/Data/ShowData/3611, https://doi.org/10.25829/idiv.3611-rgwa69).

# Load libraries

```{r}
library(dplyr)
library(tidyr)
library(readr)
library(here)
```

# Load data from Nat. Comms. paper

From https://idata.idiv.de/ddm/Data/ShowData/3611

A pan-Europe dataset of estimated linear trends in multiple dimension of local plant diversity in 57,390 vegetation-plot time series.

```{r}
biodiv_orig <- read_tsv(here("data", "3611_18_all_lm_trends_original_scale.txt"))
biodiv_perc <- read_tsv(here("data", "3611_18_all_lm_trends_percentage_change_scale.txt"))
biodiv_orig
biodiv_perc
```

# Load original data

```{r}
data_V2025.2 <- read_tsv(
  "C:/Data/MOTIVATE/MOTIVATE_data/Data/EVA and ReSurvey Data/2025 data - V2025.2/Core data/200_Motivate20250415_notJUICE_header_with_precise_coordinates.csv") %>%
  mutate(date = dmy(`Date of recording`), year = as.integer(year(date)))
```

# Check time series with many observations

```{r}
biodiv_orig %>% dplyr::filter(n == 252) %>% pull(time_seriesID)
```

```{r}
data_V2025.2 %>% dplyr::filter(`ReSurvey project`== "Biochange_meadows" & 
                                 `ReSurvey site` =="EL" & `ReSurvey plot` == "EL")
```

```{r}
data_V2025.2 %>% 
  dplyr::filter(`ReSurvey project`== "Biochange_meadows" & 
                  `ReSurvey site` =="EL" & `ReSurvey plot` == "EL") %>%
  count(year, date)
```

```{r}
data_V2025.2 %>% 
  mutate(time_seriesID = paste(`ReSurvey project`, `ReSurvey site`, `ReSurvey plot`, sep = "-")) %>% 
  group_by(time_seriesID, year) %>%
  # For each time_seriesID and year, how many different habitat assignments?
  summarise(n_different_ESy_assignments = n_distinct(`Expert system`), .groups = "drop_last") %>%
  ungroup() %>%
  mutate(more_than_one_ESy_assignment = ifelse(n_different_ESy_assignments > 1, TRUE, FALSE)) %>%
  # More than one habitat assigned to the time series in a year?
  count(more_than_one_ESy_assignment)
```

# Session info

```{r}
sessionInfo()
```

