---
title: "MOTIVATE mapping - Processing biodiversity data"
author: "Alicia Valdés"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: 4
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  word_document:
    toc: yes
    toc_depth: '4'
editor_options:
  chunk_output_type: inline
---

This notebook processes biodiversity data from Nature Communications paper by Stephan Kambach (https://doi.org/10.1038/s41467-026-72112-5). The data includes time series-specific linear trends in all biodiversity indices, expressed in annual percentage changes and in absolute units, and is openly available at the data repository of the German Centre for Integrated Biodiversity Research (iDiv) Halle-Jena-Leipzig (idata.idiv.de/ddm/Data/ShowData/3611, https://doi.org/10.25829/idiv.3611-rgwa69). The notebook joins this data with habitat extent data.

# Load libraries

```{r}
library(dplyr)
library(tidyr)
library(readr)
library(here)
library(lubridate)
library(sf)
```

# Load data from Nat. Comms. paper

From https://idata.idiv.de/ddm/Data/ShowData/3611

A pan-Europe dataset of estimated linear trends in multiple dimension of local plant diversity in 57,390 vegetation-plot time series.

```{r}
biodiv_perc <- read_tsv(here("data", "all_lm_trends_percentage_change_scale.txt"))
biodiv_perc
```

# Load original data

```{r}
data_V2025.2 <- read_tsv(
  "C:/Data/MOTIVATE/MOTIVATE_data/Data/EVA and ReSurvey Data/2025 data - V2025.2/Core data/200_Motivate20250415_notJUICE_header_with_precise_coordinates.csv") %>%
  mutate(date = dmy(`Date of recording`), year = as.integer(year(date)))
```

# Load data on extent changes

```{r}
extent_change_sf <- readRDS(here("data", "clean", "extent_change_sf.rds"))
```

# Read bioregions shapefile

```{r}
biogeo <- st_read("C:/Data/PAs/Biogeoregions/PAs/N2K_BiogeoUNITS.shp")
```

# Spatial join extent changes - bioregions

```{r}
# Ensure same CRS
biogeo <- st_transform(
  biogeo,
  st_crs(extent_change_sf)
)

# Spatial join
extent_change_sf <- extent_change_sf %>%
  st_join(
    biogeo %>%
      dplyr::select(BIOGEO, UNIT),
    join = st_intersects,
    left = TRUE
  )
```

# Add columns for EUNIS1 (start and end) to biodiversity data

```{r}
biodiv_perc <- biodiv_perc %>%
  mutate(EUNIS1_start = str_sub(EUNIS_start, 1, 1),
         EUNIS1_end = str_sub(EUNIS_end, 1, 1))
```

# Join extent and biodiversity data

```{r}
biodiv_extent <- extent_change_sf %>% st_drop_geometry() %>%
  # inner_join only joins records present in both
  # some are present in extent_change_sf but not in biodiv_perc (e.g. DK presence-absence data)
  # some are present in biodiv_perc but not in extent_change_sf (e.g. those not belonging to Q, R, S, T)
  inner_join(biodiv_perc, by = c("plotID" = "time_seriesID"),
            relationship = "many-to-many")
```

# Check if years match

```{r}
# Number of plots
biodiv_extent %>% distinct(plotID) %>% nrow()
# Number of plots where first and last year match --> USE THIS SO FAR!
biodiv_extent %>% filter(first_year == year_start & last_year == year_end) %>% 
  distinct(plotID) %>% nrow()
# Number of plots where first year does not match, but last year matches
biodiv_extent %>% filter(first_year != year_start & last_year == year_end) %>% 
  distinct(plotID) %>% nrow()
# Number of plots where first year matches, but last year does not
biodiv_extent %>% filter(first_year == year_start & last_year != year_end) %>% 
  distinct(plotID) %>% nrow()
# Number of plots where neither first nor last year match
biodiv_extent %>% filter(first_year != year_start & last_year != year_end) %>% 
  distinct(plotID) %>% nrow()
```

In some cases, they do not match because the first or last observation in some time series did not have data on functional and/or phylogenetic diversity metrics (cf. Stephan's email). So far, we use only time series where first and last year match.

```{r}
biodiv_extent <- biodiv_extent %>%
  filter(first_year == year_start & last_year == year_end)
```

```{r}
nrow(biodiv_extent %>% distinct(plotID))
```

# Map of plots in bioregions

Read shapefile with info on bioregions:

```{r}
biogeo <- st_read("C:/Data/PAs/Biogeoregions/PAs/N2K_BiogeoUNITS.shp")
```

```{r}
biodiv_extent_sf <- extent_change_sf %>% 
  filter(plotID %in% biodiv_extent$plotID)
```

```{r}
# --------------------------------------------------
# One location per plot
# --------------------------------------------------

plot_locations <- biodiv_extent_sf %>%
  distinct(plotID, .keep_all = TRUE) %>%
  mutate(
    BIOUNIT = paste(BIOGEO, UNIT, sep = "_")
  )

# --------------------------------------------------
# Add BIOUNIT to polygons
# --------------------------------------------------

biogeo_plot <- biogeo %>%
  mutate(
    BIOUNIT = paste(BIOGEO, UNIT, sep = "_")
  )

# --------------------------------------------------
# Sampled units
# --------------------------------------------------

sampled_units <- unique(plot_locations$BIOUNIT)

biogeo_plot <- biogeo_plot %>%
  mutate(
    sampled = BIOUNIT %in% sampled_units
  )

biogeo_sampled <- biogeo_plot %>%
  filter(sampled)

# --------------------------------------------------
# Zoom extent
# --------------------------------------------------

bb <- st_bbox(biogeo_sampled)

# --------------------------------------------------
# Base colours by BIOGEO
# --------------------------------------------------

biogeo_cols <- c(
  ALP = "#2166AC",  # blue
  ATL = "#F4A300",  # orange
  CON = "#1B9E77",  # green
  MED = "#D73027",  # red
  PAN = "#7B3294"   # purple
)

# --------------------------------------------------
# Create shades within each BIOGEO
# --------------------------------------------------

biounit_df <- plot_locations %>%
  st_drop_geometry() %>%
  distinct(BIOGEO, UNIT, BIOUNIT) %>%
  arrange(BIOGEO, UNIT)

cols <- c()

for (bg in unique(biounit_df$BIOGEO)) {

  units_bg <- biounit_df %>%
    filter(BIOGEO == bg)

  n_units <- nrow(units_bg)

  pal <- colorRampPalette(
    c(
      lighten(biogeo_cols[bg], amount = 0.35),
      darken(biogeo_cols[bg], amount = 0.15)
    )
  )(n_units)

  cols <- c(
    cols,
    setNames(pal, units_bg$BIOUNIT)
  )
}

# --------------------------------------------------
# Plot
# --------------------------------------------------

ggplot() +

  # Background units
  geom_sf(
    data = biogeo_plot,
    fill = "grey95",
    color = "grey75",
    linewidth = 0.2
  ) +

  # Sampled units
  geom_sf(
    data = biogeo_sampled,
    aes(fill = BIOUNIT),
    color = "grey40",
    linewidth = 0.3
  ) +

  # Plot locations
  geom_sf(
    data = plot_locations,
    color = "black",
    size = 0.12,
    alpha = 0.4
  ) +

  scale_fill_manual(
    values = cols,
    name = "Biogeographical unit"
  ) +

  coord_sf(
    xlim = c(bb["xmin"], bb["xmax"]),
    ylim = c(bb["ymin"], bb["ymax"]),
    expand = FALSE
  ) +

  labs(
    title = "Distribution of plots across biogeographical units"
  ) +

  theme_bw() +

  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title = element_blank(),
    legend.position = "right",
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 8),
    plot.title = element_text(
      face = "bold",
      hjust = 0.5
    )
  )
```

# Count of plots in bioregions

```{r}
biodiv_extent_sf %>% st_drop_geometry() %>% select(plotID, BIOGEO, UNIT) %>% 
  distinct() %>% count(BIOGEO) %>% arrange(n)
biodiv_extent_sf %>% st_drop_geometry() %>% select(plotID, BIOGEO, UNIT) %>% 
  distinct() %>% count(BIOGEO, UNIT) %>% arrange(n)
```

# Add geometry

```{r}
biodiv_extent <- biodiv_extent %>%
  left_join(biodiv_extent_sf %>% distinct(plotID, geometry),
    by = c("plotID"),
    relationship = "many-to-one"
  ) %>%
  st_as_sf()
```

## Save as .rds

```{r}
saveRDS(biodiv_extent, here("data", "clean", "biodiv_extent.rds"))
```

# Session info

```{r}
sessionInfo()
```

