---
title: "MOTIVATE mapping - Analysing biodiversity - habitat exent relationships"
author: "Alicia Valdés"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: 4
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  word_document:
    toc: yes
    toc_depth: '4'
editor_options:
  chunk_output_type: inline
---

This notebook analyses relationships between biodiversity metrics and habitat extent metrics.

# Load libraries

```{r}
library(dplyr)
library(tidyr)
library(readr)
library(here)
library(purrr)
library(broom)
library(MASS)
library(ggplot2)
library(corrplot)
```

# Load data on biodiversity and habitat extent

```{r}
biodiv_extent <- readRDS(here("data", "clean", "biodiv_extent.rds"))
```


```{r}
biodiv_extent <- biodiv_extent %>%
  sf::st_drop_geometry() %>%
  pivot_longer(cols = absolute_change_perc:relative_change,
               names_to = "metric_extent", values_to = "value") %>% 
  rename(metric_biodiv = metric)
```

# Distributions

## Distributions of initial biodiversity values

```{r}
# Histograms
biodiv_extent %>%
  group_by(metric_biodiv) %>%
  mutate(
    q01 = quantile(initial_div_value, 0.01, na.rm = TRUE),
    q99 = quantile(initial_div_value, 0.99, na.rm = TRUE)
  ) %>%
  filter(initial_div_value >= q01, initial_div_value <= q99) %>%
  ggplot(aes(initial_div_value)) +
  geom_histogram(bins = 30, color = "black", fill = "grey") +
  facet_wrap(~metric_biodiv, scales = "free", ncol = 6) +
  theme_bw() + xlab("Initial diversity value")
# Density plots
biodiv_extent %>%
  group_by(metric_biodiv) %>%
  mutate(
    q01 = quantile(initial_div_value, 0.01, na.rm = TRUE),
    q99 = quantile(initial_div_value, 0.99, na.rm = TRUE)
  ) %>%
  filter(initial_div_value >= q01, initial_div_value <= q99) %>%
  ggplot(aes(initial_div_value)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  facet_wrap(~metric_biodiv, scales = "free")
# Boxplots
ggplot(biodiv_extent,
       aes(x = metric_biodiv, y = initial_div_value)) +
  geom_boxplot(outlier.alpha = 0.1) +
  coord_flip()
```

## Distributions of biodiversity change

```{r}
# Histograms
biodiv_extent %>%
  group_by(metric_biodiv) %>%
  mutate(
    q01 = quantile(slope, 0.01, na.rm = TRUE),
    q99 = quantile(slope, 0.99, na.rm = TRUE)
  ) %>%
  filter(slope >= q01, slope <= q99) %>%
  ggplot(aes(slope)) +
  geom_histogram(bins = 30, color = "black", fill = "grey") +
  facet_wrap(~metric_biodiv, scales = "free", ncol = 6) +
  theme_bw() + xlab("Regression slope (annual percentage change)")
# Density plots
biodiv_extent %>%
  group_by(metric_biodiv) %>%
  mutate(
    q01 = quantile(slope, 0.01, na.rm = TRUE),
    q99 = quantile(slope, 0.99, na.rm = TRUE)
  ) %>%
  filter(slope >= q01, slope <= q99) %>%
  ggplot(aes(slope)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  facet_wrap(~metric_biodiv, scales = "free")
# Boxplots
ggplot(biodiv_extent,
       aes(x = metric_biodiv, y = slope)) +
  geom_boxplot(outlier.alpha = 0.1) +
  coord_flip()
```

## Distributions of habitat extent change

```{r}
# Histograms
biodiv_extent %>%
  group_by(metric_extent, percentile) %>%
  mutate(
    q01 = quantile(value, 0.01, na.rm = TRUE),
    q99 = quantile(value, 0.99, na.rm = TRUE)
  ) %>%
  filter(value >= q01, value <= q99) %>%
  ggplot(aes(value)) +
  geom_histogram(bins = 30, color = "black", fill = "grey") +
  facet_wrap(~metric_extent + percentile, scales = "free") +
  theme_bw() + xlab("Change in extent")
# Density plots
biodiv_extent %>%
  group_by(metric_extent, percentile) %>%
  mutate(
    q01 = quantile(value, 0.01, na.rm = TRUE),
    q99 = quantile(value, 0.99, na.rm = TRUE)
  ) %>%
  filter(value >= q01, value <= q99) %>%
  ggplot(aes(value)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  facet_wrap(~metric_extent + percentile, scales = "free")
```

# Correlations

## Correlations among initial biodiversity values

```{r}
biodiv_wide <- biodiv_extent %>%
  distinct(plotID, metric_biodiv, initial_div_value) %>%
  pivot_wider(names_from = metric_biodiv, values_from = initial_div_value)
```

```{r}
cor_mat_init <- biodiv_wide %>%
  dplyr::select(-plotID) %>%
  cor(use = "pairwise.complete.obs")
corrplot(cor_mat_init, method = "color", type = "upper", addCoef.col = "black",
         number.cex = 0.6, tl.col = "black", order = "hclust")
```

## Correlations among biodiversity change metrics

```{r}
biodiv_change_wide <- biodiv_extent %>%
  distinct(plotID, metric_biodiv, slope) %>%
  pivot_wider(names_from = metric_biodiv, values_from = slope)
```

```{r}
cor_mat_change <- biodiv_change_wide %>%
  dplyr::select(-plotID) %>%
  cor(use = "pairwise.complete.obs")
corrplot(cor_mat_change, method = "color", type = "upper", addCoef.col = "black",
         number.cex = 0.6, tl.col = "black", order = "hclust")
```

# Models

## Effect of initial extent on initial diversity

```{r}
biodiv_extent_first_ha <- biodiv_extent %>%
  distinct(plotID, percentile, EUNIS_plot, extent_first_ha, metric_biodiv, 
           initial_div_value, Trajectory) %>%
  # Remove the bottom and top 1% of extent_first_ha
  filter(between(extent_first_ha,
                 quantile(extent_first_ha, 0.01),
                 quantile(extent_first_ha, 0.99)))
```

```{r}
models_init <- biodiv_extent_first_ha %>%
  group_by(metric_biodiv, percentile) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(initial_div_value ~ extent_first_ha, data = .x)),
    coef = map(model, ~ tidy(.x) %>%
                 filter(term == "extent_first_ha")),
    fit = map(model, glance)
  ) %>%
  transmute(
    metric_biodiv,
    percentile,
    estimate = map_dbl(coef, "estimate"),
    p.value = map_dbl(coef, "p.value"),
    r2 = map_dbl(fit, "r.squared"),
    n = map_int(data, nrow)
  ) %>%
  arrange(desc(r2))
```

```{r}
models_init %>% filter(p.value < 0.1 & r2 > 0.01)
```
```{r}
models_init %>% filter(p.value < 0.05) %>% arrange(metric_biodiv, percentile)
```

```{r}
plot_data_init <- biodiv_extent_first_ha %>%
  semi_join(
    models_init %>%
      filter(p.value < 0.05) %>%
      dplyr::select(metric_biodiv, percentile),
    by = c("metric_biodiv", "percentile")
  ) %>%
  left_join(
    models_init %>%
      dplyr::select(metric_biodiv, percentile, estimate, p.value, r2, n),
    by = c("metric_biodiv", "percentile")
  )
```

```{r}
make_plot_init <- function(df){

  biodiv_name <- unique(df$metric_biodiv)
  perc <- unique(df$percentile)
  estimate <- unique(df$estimate)
  pval <- unique(df$p.value)
  r2 <- unique(df$r2)
  n <- unique(df$n)

  ggplot(df, aes(x = extent_first_ha, y = initial_div_value)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "lm", colour = "blue", fill = "blue", alpha = 0.15) +
    geom_smooth(method = MASS::rlm, colour = "red", se = FALSE) +
    annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.5, size = 4,
             label = sprintf("β = %.3f\nR² = %.3f\np = %.3g\nn = %d",
                             estimate, r2, pval, n)) +
    labs(title = paste0(biodiv_name, " | percentile = ", perc),
         x = "Initial habitat extent (ha)",
         y = paste0("Initial value of ", biodiv_name), colour = NULL) +
    theme_bw()
}
```

```{r message=FALSE}
plot_data_init %>%
  group_by(metric_biodiv, percentile) %>%
  group_split() %>%
  map(make_plot_init)
```

## Effect of change in extent on change in diversity

```{r}
biodiv_absolute_change_ha <- biodiv_extent %>%
  filter(metric_extent == "absolute_change_ha") %>%
  distinct(plotID, percentile, EUNIS_plot, metric_biodiv, slope, 
           Trajectory, value) %>%
  rename(absolute_change_ha = value) %>%
  # Remove the bottom and top 1% of absolute_change_ha
  filter(between(absolute_change_ha,
                 quantile(absolute_change_ha, 0.01),
                 quantile(absolute_change_ha, 0.99)))
```

```{r}
models_change <- biodiv_absolute_change_ha %>%
  group_by(metric_biodiv, percentile) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(slope ~ absolute_change_ha, data = .x)),
    coef = map(model, ~ tidy(.x) %>%
                 filter(term == "absolute_change_ha")),
    fit = map(model, glance)
  ) %>%
  transmute(
    metric_biodiv,
    percentile,
    estimate = map_dbl(coef, "estimate"),
    p.value = map_dbl(coef, "p.value"),
    r2 = map_dbl(fit, "r.squared"),
    n = map_int(data, nrow)
  ) %>%
  arrange(desc(r2))
```

```{r}
models_change %>% filter(p.value < 0.1 & r2 > 0.01)
```

```{r}
models_change %>% filter(p.value < 0.1) %>% arrange(metric_biodiv, percentile)
```

```{r}
plot_data_change <- biodiv_absolute_change_ha %>%
  semi_join(
    models_change %>%
      filter(p.value < 0.1) %>%
      dplyr::select(metric_biodiv, percentile),
    by = c("metric_biodiv", "percentile")
  ) %>%
  left_join(
    models_change %>%
      dplyr::select(metric_biodiv, percentile, estimate, p.value, r2, n),
    by = c("metric_biodiv", "percentile")
  )
```

```{r}
make_plot_change <- function(df){

  biodiv_name <- unique(df$metric_biodiv)
  perc <- unique(df$percentile)
  estimate <- unique(df$estimate)
  pval <- unique(df$p.value)
  r2 <- unique(df$r2)
  n <- unique(df$n)

  ggplot(df, aes(x = absolute_change_ha, y = slope)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "lm", colour = "blue", fill = "blue", alpha = 0.15) +
    geom_smooth(method = MASS::rlm, colour = "red", se = FALSE) +
    annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.5, size = 4,
             label = sprintf("β = %.3f\nR² = %.3f\np = %.3g\nn = %d",
                             estimate, r2, pval, n)) +
    labs(title = paste0(biodiv_name, " | percentile = ", perc), 
         x = "Absolute change in extent (ha)",
         y = paste0("Change in ", biodiv_name), colour = NULL) +
    theme_bw()
}
```

```{r message=FALSE}
plot_data_change %>%
  group_by(metric_biodiv, percentile) %>%
  group_split() %>%
  map(make_plot_change)
```

## Effect of change in extent for each habitat type

```{r}
models_change_hab <- biodiv_absolute_change_ha %>%
  group_by(metric_biodiv, percentile, EUNIS_plot) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(slope ~ absolute_change_ha, data = .x)),
    coef = map(model, ~ tidy(.x) %>%
                 filter(term == "absolute_change_ha")),
    fit = map(model, glance)
  ) %>%
  transmute(
    metric_biodiv,
    percentile,
    EUNIS_plot,
    estimate = map_dbl(coef, "estimate"),
    p.value = map_dbl(coef, "p.value"),
    r2 = map_dbl(fit, "r.squared"),
    n = map_int(data, nrow)
  ) %>%
  arrange(desc(r2))
```

```{r}
models_change_hab %>% filter(p.value < 0.05) %>% arrange(metric_biodiv, percentile)
```

```{r}
plot_data_change_hab <- biodiv_absolute_change_ha %>%
  semi_join(
    models_change_hab %>%
      dplyr::filter(p.value < 0.05),
    by = c("metric_biodiv", "percentile", "EUNIS_plot")
  ) %>%
  left_join(
    models_change_hab %>%
      dplyr::select(metric_biodiv, percentile, EUNIS_plot, estimate, p.value, r2, n),
    by = c("metric_biodiv", "percentile", "EUNIS_plot")
  )
```

```{r}
make_plot_change_hab <- function(df){

  biodiv_name <- unique(df$metric_biodiv)
  perc <- unique(df$percentile)
  EUNIS <- unique(df$EUNIS_plot)
  estimate <- unique(df$estimate)
  pval <- unique(df$p.value)
  r2 <- unique(df$r2)
  n <- unique(df$n)

  ggplot(df, aes(x = absolute_change_ha, y = slope)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "lm", colour = "blue", fill = "blue", alpha = 0.15) +
    geom_smooth(method = MASS::rlm, colour = "red", se = FALSE) +
    annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.5, size = 4,
             label = sprintf("β = %.3f\nR² = %.3f\np = %.3g\nn = %d",
                             estimate, r2, pval, n)) +
    labs(title = paste0(biodiv_name, " | percentile = ", perc,
                        " | EUNIS = ", EUNIS),
         x = "Absolute change in extent (ha)",
         y = paste0("Change in ", biodiv_name)) +
    theme_bw()
  }
```

```{r message=FALSE}
plot_data_change_hab %>%
  group_by(metric_biodiv, percentile, EUNIS_plot) %>%
  group_split() %>%
  map(make_plot_change_hab)
```

## Effect of change in extent for each trajectory type

```{r}
models_change_traj <- biodiv_absolute_change_ha %>%
  group_by(metric_biodiv, percentile, Trajectory) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(slope ~ absolute_change_ha, data = .x)),
    coef = map(model, ~ tidy(.x) %>%
                 filter(term == "absolute_change_ha")),
    fit = map(model, glance)
  ) %>%
  transmute(
    metric_biodiv,
    percentile,
    Trajectory,
    estimate = map_dbl(coef, "estimate"),
    p.value = map_dbl(coef, "p.value"),
    r2 = map_dbl(fit, "r.squared"),
    n = map_int(data, nrow)
  ) %>%
  arrange(desc(r2))
```

```{r}
models_change_traj %>% filter(p.value < 0.05) %>% arrange(metric_biodiv, percentile)
```

```{r}
plot_data_change_traj <- biodiv_absolute_change_ha %>%
  semi_join(
    models_change_traj %>%
      dplyr::filter(p.value < 0.1 & r2 > 0.01) %>%
      dplyr::select(metric_biodiv, percentile, Trajectory),
    by = c("metric_biodiv", "percentile", "Trajectory")
  ) %>%
  left_join(
    models_change_traj %>%
      dplyr::select(metric_biodiv, percentile, Trajectory, estimate, p.value, r2, n),
    by = c("metric_biodiv", "percentile", "Trajectory")
  )
```

```{r}
make_plot_change_traj <- function(df){

  biodiv_name <- unique(df$metric_biodiv)
  perc <- unique(df$percentile)
  traj <- unique(df$Trajectory)
  estimate <- unique(df$estimate)
  pval <- unique(df$p.value)
  r2 <- unique(df$r2)
  n <- unique(df$n)

  ggplot(df, aes(x = absolute_change_ha, y = slope)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "lm", colour = "blue", fill = "blue", alpha = 0.15) +
    geom_smooth(method = MASS::rlm, colour = "red", se = FALSE) +
    annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.5, size = 4,
             label = sprintf("β = %.3f\nR² = %.3f\np = %.3g\nn = %d",
                             estimate, r2, pval, n)) +
    labs(title = paste0(biodiv_name, " | percentile = ", perc, 
                        " | Trajectory = ", traj),
         x = "Absolute change in extent (ha)", 
         y = paste0("Change in ", biodiv_name)) +
    theme_bw()
  }
```

```{r message=FALSE}
plot_data_change_traj %>%
  group_by(metric_biodiv, percentile, Trajectory) %>%
  group_split() %>%
  map(make_plot_change_traj)
```

# IDEA

Quantify extent change by using regression slopes between the observation year and the habitat extent measure (more similar to biodiversity changes in Nat. Comms. paper).

# Session info

```{r}
sessionInfo()
```

