---
title: "MOTIVATE mapping - Data on extent and functional trends from RF"
author: "Alicia Valdés"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: 4
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  word_document:
    toc: yes
    toc_depth: '4'
editor_options:
  chunk_output_type: inline
---

This notebook analyses data on habitat extent metrics and functional trends derived from Random Forest models.

# Load libraries

```{r}
library(dplyr)
library(tidyr)
library(readr)
library(here)
library(corrplot)
library(ggplot2)
library(gghalves)
library(car)
library(emmeans)
library(multcomp)
library(ggeffects)
library(patchwork)
library(stringr)
```

# Load data

```{r}
# Data of plots with constant habitat
RF_data_nochange <- read_csv(here("data", "RF", "MOTIVATE_ALL_V1_SPECTRAL.csv"))
# Data of plots with habitat change
RF_data_change <- read_csv(here("data", "RF", "MOTIVATE_CHANGE_ALL_V1_SPECTRAL.csv"))
```

# Load initial dataset from validation

Initial dataset with info on EUNIS 2.

```{r}
initial_dataset <- read_csv(here("..", "MOTIVATE_validation", "data", "clean",
                                 "ReSurvey_QRST_data_V2025.2_20261007.csv"))
```

# Remove plots with habitat change from dataset of plots with constant habitat

```{r}
RF_data_nochange <- RF_data_nochange %>%
  filter(!plot_id %in% RF_data_change$plot_id)
```

# HERE: Check some columns that seem to be duplicated

```{r}
RF_data_nochange  %>% filter(coalesce(eunis1, "NA") != coalesce(EUNIS1, "NA"))
RF_data_nochange  %>% filter(is.na(eunis1))
RF_data_nochange  %>% filter(coalesce(biogeo, "NA") != coalesce(BIOGEO, "NA"))
RF_data_nochange  %>% filter(is.na(biogeo))
RF_data_nochange  %>% filter(first_year.x != first_year.y) # 3 plots
RF_data_nochange  %>% filter(last_year.x != last_year.y)
RF_data_nochange  %>% filter(years_between.x != years_between.y) # 3 plots
```

Check with my data which first_year is correct!

```{r}
RF_data_change  %>% filter(coalesce(biogeo, "NA") != coalesce(BIOGEO, "NA"))
RF_data_change %>% filter(is.na(biogeo))
RF_data_change  %>% filter(first_year.x != first_year.y)
RF_data_change  %>% filter(last_year.x != last_year.y)
RF_data_change  %>% filter(years_between.x != years_between.y)
```

## Remove duplicated columns and rename remaining ones

```{r}
RF_data_nochange <- RF_data_nochange %>%
  # TBD for first_year and years_between!
  dplyr::select(-EUNIS1, -BIOGEO, -last_year.y) %>%
  rename(last_year = last_year.x)
```

```{r}
RF_data_change <- RF_data_change %>%
  dplyr::select(-BIOGEO)
```

## Join plots with constant habitat and plots with habitat change

```{r}
RF_data_change <- RF_data_change %>% 
  dplyr::select(-EUNIS1, -last_year.y) %>%
  # Set eunis1 as "Change" for plots with habitat change
  mutate(eunis1 = "Change") %>%
  rename(last_year = last_year.x)
```


```{r}
RF_data <- bind_rows(RF_data_nochange, RF_data_change)
```

## Fix plotIDs in RF_data

Gonzalo used RS_CODE instead of Resurvey project as first part of plot_id.

Retrieve list of correspondences of RS_CODE and Resurvey project from original data:

```{r}
correspondence_list <- initial_dataset %>%
  distinct(RS_CODE, `ReSurvey project`)
```

```{r}
RF_data <- RF_data %>%
  # Add spaces before and after "-"
  mutate(plotID = str_replace_all(plot_id, "-", " - ")) %>%
  # Get RS_CODE: first part of plot_id, before first "-"
  mutate(RS_CODE = sub("-.*", "", plot_id)) %>%
  # Join to get Resurvey project
  left_join(correspondence_list) %>%
  # Construct new plotID
  mutate(plotID = str_replace(plotID, paste0("^", RS_CODE), 
                              `ReSurvey project`)) %>%
  # Remove unneeded columns
  dplyr::select(-plot_id, -RS_CODE, -`ReSurvey project`) %>%
  # Modify some plotIDs
  mutate(
    plotID = str_replace(
      plotID,
      "^AOPK_HabitatMonitoring - PolickeSteny - Signal -",
      "AOPK_HabitatMonitoring - PolickeSteny-Signal -"
      ),
    plotID = str_replace(
      plotID,
      "^Ecohydrology of Floodplain Meadows - ([0-9]+) - \\1\\s+-\\s+",
      "Ecohydrology of Floodplain Meadows - \\1 - \\1-"
      ),
    plotID = str_replace(
      plotID,
      "^PSET Puy Long - PuyLong - PSET 0 - 1",
      "PSET Puy Long - PuyLong - PSET 0-1"
      ),
    plotID = str_replace(
      plotID,
      "^PSET Puy Long - PuyLong - PSET ([0-9]+) - ([0-9]+)",
      "PSET Puy Long - PuyLong - PSET \\1-\\2"
      ),
    plotID = str_replace(
      plotID,
      "^Role of AES schemes on the grasslands biodiversity - ([0-9]+) - ([0-9]+) - ([0-9]+) - \\1 - \\2 - \\3(_[0-9]+)?$",
      "Role of AES schemes on the grasslands biodiversity - \\1-\\2-\\3 - \\1-\\2-\\3\\4"
      )
    )
```

Two types of plot IDs:

```{r}
RF_data <- RF_data %>%
  # Without spaces before and after "-"
  mutate(plot_id = str_replace_all(plotID, " - ", "-")) %>%
  # With spaces before and after "-"
  rename(plot_id_spaced = plotID)
```

## Add EUNIS level 2 

Useful for some analyses.

```{r}
data_eunis <- initial_dataset %>% 
  dplyr::select(PlotObservationID, `ReSurvey project`, RS_CODE, `ReSurvey site`, 
                `ReSurvey plot`, year, plotID_AV, EUNIS1, EUNIS2) %>%
  # Remove spaces in plot_id
  mutate(plot_id = str_replace_all(plotID_AV, " - ", "-")) %>% 
  # Remove unneeded columns
  dplyr::select(-PlotObservationID, -`ReSurvey project`, -RS_CODE, 
                -`ReSurvey site`, -`ReSurvey plot`, -plotID_AV) %>%
  # Keep only plot IDs in RF_data
  filter(plot_id %in% RF_data$plot_id) %>% 
  # To check match with EUNIS 1 in RF data
  rename(EUNIS1_check = EUNIS1)
```

```{r}
RF_data <- RF_data %>%
  left_join(data_eunis %>% rename(first_year.x = year, 
                                  EUNIS1_check_first = EUNIS1_check,
                                  eunis2_first = EUNIS2),
            by = c("plot_id", "first_year.x")) %>%
  left_join(data_eunis %>% rename(last_year = year, 
                                  EUNIS1_check_last = EUNIS1_check,
                                  eunis2_last = EUNIS2),
            by = c("plot_id", "last_year"))
```

Some checks:

```{r}
RF_data %>% filter(eunis_cambio == F & eunis1 != EUNIS1_check_first)
RF_data %>% filter(eunis_cambio == F & eunis1 != EUNIS1_check_last)
RF_data %>% filter(eunis_cambio == T & eunis_first != EUNIS1_check_first)
RF_data %>% filter(eunis_cambio == T & eunis_last != EUNIS1_check_last)
```

All OK.

```{r}
RF_data <- RF_data %>%
  # Remove unneeded columns
  dplyr::select(-EUNIS1_check_first, -EUNIS1_check_last) %>%
  # create eunis2 column
  mutate(eunis2 = case_when(
    eunis1 == "Change" ~ "Change",
    eunis2_first == eunis2_last ~ eunis2_first,
    !is.na(eunis2_first) & is.na(eunis2_last) ~ eunis2_first,
    is.na(eunis2_first) & !is.na(eunis2_last) ~ eunis2_last,
    eunis2_first != eunis2_last ~ "Change"
    ))
```

## Save as .rds

```{r}
saveRDS(RF_data, here("data", "clean", "RF_data.rds"))
```

# Change in extent

## Plots with constant habitat

### Change in extent by habitat type 

```{r}
mod_extent_eunis1_noch <- lm(delta_focal_pct ~ eunis1, 
                             data = RF_data %>% filter(eunis_cambio == F))
Anova(mod_extent_eunis1_noch)
```

```{r}
emmeans(mod_extent_eunis1_noch, pairwise ~ eunis1)
```

Proportion of variance explained:

```{r}
summary(mod_extent_eunis1_noch)$r.squared
```

### Change in extent by number of years between first and last survey

```{r}
mod_extent_yrs_eunis1_noch <- lm(abs(delta_focal_pct) ~ years_between.x * eunis1,
          data = RF_data %>% filter(eunis_cambio == F))
Anova(mod_extent_yrs_eunis1_noch)
summary(mod_extent_yrs_eunis1_noch)
```

```{r}
emtrends(mod_extent_yrs_eunis1_noch, pairwise ~ eunis1, var = "years_between.x")
```

Proportion of variance explained:

```{r}
summary(mod_extent_yrs_eunis1_noch)$r.squared
```

### Figure 1

```{r}
fig1a <- ggplot(RF_data %>% filter(eunis_cambio == F), 
                aes(x = eunis1, y = delta_focal_pct, fill = eunis1)) +
  #geom_half_violin(side = "r", alpha = 0.8, position = position_nudge(x = 0.2)) +
  geom_point(aes(color = eunis1),
             position = position_jitter(width = 0.25), size = 0.5, alpha = 0.1) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2,
                linewidth = 0.3, color = "black") +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 2,
                color = "black", fill = "black") +
  # geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.8) +
  # geom_text(data = cld(emmeans(mod_extent_eunis1_noch, ~ eunis1), 
  #                      Letters = letters) %>%
  #             as.data.frame(),
  #           aes(x = eunis1, y = 75, label = .group),
  #           inherit.aes = FALSE, size = 5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "EUNIS I Habitat type", y = "Change in habitat extent (%)") +
  theme_bw() +
  theme(legend.position = "none") 
fig1a
```

```{r}
fig1b <- ggplot(RF_data %>% filter(eunis_cambio == F),
                aes(x = years_between.x, y = abs(delta_focal_pct), 
                    color = eunis1, fill = eunis1, group = eunis1)) +
  # geom_point(size = 1, alpha = 0.1) +
  geom_smooth(method = "lm") +
  labs(x = "Number of years between first and last survey",
       y = "Absolute change in habitat extent (%)") +
  theme_bw() +
  theme(legend.position = "none")
fig1b
```

```{r}
fig1b_alt <- ggplot(RF_data %>% filter(eunis_cambio == F),
                aes(x = years_between.x, y = delta_focal_pct, 
                    color = eunis1, fill = eunis1, group = eunis1)) +
  # geom_point(size = 1, alpha = 0.1) +
  geom_smooth(method = "lm") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "Number of years between first and last survey",
       y = "Change in habitat extent (%)") +
  theme_bw() +
  theme(legend.position = "none")
fig1b_alt
```

```{r}
fig1 <- fig1a + fig1b_alt +
  plot_layout(ncol = 2) + plot_annotation(tag_levels = list(c("A)", "B)")))
fig1
ggsave(fig1, filename = "output/figures/fig1.jpeg", 
       width = 24, height = 12, units = "cm", dpi = 300)
```

## Plots with habitat change

### Change in extent by habitat type 

```{r}
mod_extent_eunis_first_ch <- lm(delta_focal_pct ~ eunis_first,
                                data = RF_data %>% filter(eunis_cambio == T))
Anova(mod_extent_eunis_first_ch)
```

```{r}
emmeans(mod_extent_eunis_first_ch, pairwise ~ eunis_first)
```

Proportion of variance explained:

```{r}
summary(mod_extent_eunis_first_ch)$r.squared
```

### Change in extent by transition type

```{r}
ggplot(RF_data %>% filter(eunis_cambio == T) %>%
         left_join(RF_data %>% filter(eunis_cambio == TRUE) %>%
                     count(transition),
                   by = "transition") %>%
         mutate(transition_label = paste0(transition, "\n(n = ", n, ")")),
       aes(x = transition_label, y = delta_focal_pct, fill = eunis_first)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_point(aes(color = eunis_first),
             position = position_jitter(width = 0.25), size = 2, alpha = 0.2) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2,
               linewidth = 0.3, color = "black") +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 2,
                color = "black", fill = "black") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  labs(x = "Transition", y = "Change in habitat extent (%)") +
  theme_bw() +
  theme(legend.position = "none") 
```

### Change in extent by number of years between first and last survey

```{r}
mod_extent_yrs_eunis_first_ch <- lm(abs(delta_focal_pct) ~ 
                                 years_between.x * eunis_first,
          data = RF_data %>% filter(eunis_cambio == T))
Anova(mod_extent_yrs_eunis_first_ch)
summary(mod_extent_yrs_eunis_first_ch)
```

```{r}
mod_extent_yrs_ch <- lm(abs(delta_focal_pct) ~ years_between.x ,
                        data = RF_data %>% filter(eunis_cambio == T))
Anova(mod_extent_yrs_ch)
summary(mod_extent_yrs_ch)
```

Proportion of variance explained:

```{r}
summary(mod_extent_yrs_eunis_first_ch)$r.squared
```

### Appendix 1

```{r}
app1 <- ggplot(RF_data %>% filter(eunis_cambio == T),
                aes(x = eunis_first, y = delta_focal_pct, fill = eunis_first)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  #geom_half_violin(side = "r", alpha = 0.8, position = position_nudge(x = 0.2)) +
  geom_point(aes(color = eunis_first),
             position = position_jitter(width = 0.25), size = 2, alpha = 0.2) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2,
               linewidth = 0.3, color = "black") +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 2,
                color = "black", fill = "black") +
  # geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.8) +
  # geom_text(data = cld(emmeans(mod_extent_eunis1_ch, ~ eunis_first), 
  #                      Letters = letters) %>%
  #             as.data.frame(),
  #           aes(x = eunis_first, y = 75, label = .group),
  #           inherit.aes = FALSE, size = 5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  labs(x = "Original EUNIS I Habitat type", y = "Change in habitat extent (%)") +
  theme_bw() +
  theme(legend.position = "none") 
app1
ggsave(app1, filename = "output/figures/app1.jpeg", 
       width = 12, height = 12, units = "cm", dpi = 300)
```

# Change in extent by bioregion

## Plots with constant habitat

```{r}
table((RF_data %>% filter(eunis_cambio == F))$eunis1, 
      (RF_data %>% filter(eunis_cambio == F))$biogeo)
```

### Change in extent by habitat type 

```{r}
mod_extent_eunis1_noch_ALP <- lm(delta_focal_pct ~ eunis1,
                                 data = RF_data %>% filter(eunis_cambio == F) %>%
                                   filter(biogeo == "ALP"))
mod_extent_eunis1_noch_ATL <- lm(delta_focal_pct ~ eunis1,
                                 data = RF_data %>% filter(eunis_cambio == F) %>%
                                   filter(biogeo == "ATL"))
mod_extent_eunis1_noch_CON <- lm(delta_focal_pct ~ eunis1,
                                 data = RF_data %>% filter(eunis_cambio == F) %>%
                                   filter(biogeo == "CON"))
# Not possible in MED because all are R
mod_extent_eunis1_noch_PAN <- lm(delta_focal_pct ~ eunis1,
                                 data = RF_data %>% filter(eunis_cambio == F) %>%
                                   filter(biogeo == "PAN"))
Anova(mod_extent_eunis1_noch_ALP)
Anova(mod_extent_eunis1_noch_ATL)
Anova(mod_extent_eunis1_noch_CON)
Anova(mod_extent_eunis1_noch_PAN)
```

```{r}
emmeans(mod_extent_eunis1_noch_ATL, pairwise ~ eunis1)
emmeans(mod_extent_eunis1_noch_CON, pairwise ~ eunis1)
```

Proportion of variance explained:

```{r}
summary(mod_extent_eunis1_noch_ALP)$r.squared
summary(mod_extent_eunis1_noch_ATL)$r.squared
summary(mod_extent_eunis1_noch_CON)$r.squared
summary(mod_extent_eunis1_noch_PAN)$r.squared
```

### Change in extent by number of years between first and last survey

```{r}
mod_extent_yrs_noch_ALP <- lm(abs(delta_focal_pct) ~ years_between.x,
                                     data = RF_data %>% 
                                filter(eunis_cambio == F) %>%
                                       filter(biogeo == "ALP"))
mod_extent_yrs_noch_ATL <- lm(abs(delta_focal_pct) ~ years_between.x,
                                     data = RF_data %>% 
                                filter(eunis_cambio == F) %>%
                                       filter(biogeo == "ATL"))
mod_extent_yrs_noch_CON <- lm(abs(delta_focal_pct) ~ years_between.x,
                                     data = RF_data %>% 
                                filter(eunis_cambio == F) %>%
                                       filter(biogeo == "CON"))
mod_extent_yrs_noch_MED <- lm(abs(delta_focal_pct) ~ years_between.x,
                                     data = RF_data %>% 
                                filter(eunis_cambio == F) %>%
                                       filter(biogeo == "MED"))
mod_extent_yrs_noch_PAN <- lm(abs(delta_focal_pct) ~ years_between.x,
                                     data = RF_data %>% 
                                filter(eunis_cambio == F) %>%
                                       filter(biogeo == "PAN"))
Anova(mod_extent_yrs_noch_ALP)
Anova(mod_extent_yrs_noch_ATL)
Anova(mod_extent_yrs_noch_CON)
Anova(mod_extent_yrs_noch_MED)
Anova(mod_extent_yrs_noch_PAN)
summary(mod_extent_yrs_noch_ALP)
summary(mod_extent_yrs_noch_ATL)
summary(mod_extent_yrs_noch_CON)
summary(mod_extent_yrs_noch_MED)
summary(mod_extent_yrs_noch_PAN)
```

## Plots with habitat change

```{r}
table((RF_data %>% filter(eunis_cambio == T))$eunis_first, 
      (RF_data %>% filter(eunis_cambio == T))$biogeo)
```

### Change in extent by habitat type 

```{r}
mod_extent_eunis_first_ch_ALP <- lm(delta_focal_pct ~ eunis_first, 
                                    data = RF_data %>% 
                                      filter(eunis_cambio == T) %>% 
                                      filter(biogeo == "ALP"))
mod_extent_eunis_first_ch_ATL <- lm(delta_focal_pct ~ eunis_first, 
                                    data = RF_data %>% 
                                      filter(eunis_cambio == T) %>% 
                                      filter(biogeo == "ATL"))
mod_extent_eunis_first_ch_CON <- lm(delta_focal_pct ~ eunis_first, 
                                    data = RF_data %>% 
                                      filter(eunis_cambio == T) %>% 
                                      filter(biogeo == "CON"))
Anova(mod_extent_eunis_first_ch_ALP)
Anova(mod_extent_eunis_first_ch_ATL)
Anova(mod_extent_eunis_first_ch_CON)
```

```{r}
emmeans(mod_extent_eunis_first_ch_ATL, pairwise ~ eunis_first)
```

Proportion of variance explained:

```{r}
summary(mod_extent_eunis_first_ch_ALP)$r.squared
summary(mod_extent_eunis_first_ch_ATL)$r.squared
summary(mod_extent_eunis_first_ch_CON)$r.squared
```

### Change in extent by number of years between first and last survey

```{r}
mod_extent_yrs_ch_ALP <- lm(abs(delta_focal_pct) ~ years_between.x,
          data = RF_data %>% filter(eunis_cambio == T) %>% 
            filter(biogeo == "ALP"))
mod_extent_yrs_ch_ATL <- lm(abs(delta_focal_pct) ~ years_between.x,
          data = RF_data %>% filter(eunis_cambio == T) %>% 
            filter(biogeo == "ATL"))
mod_extent_yrs_ch_CON <- lm(abs(delta_focal_pct) ~ years_between.x,
          data = RF_data %>% filter(eunis_cambio == T) %>% 
            filter(biogeo == "CON"))
Anova(mod_extent_yrs_ch_ALP)
Anova(mod_extent_yrs_ch_ATL)
Anova(mod_extent_yrs_ch_CON)
summary(mod_extent_yrs_ch_ALP)
summary(mod_extent_yrs_ch_ATL)
summary(mod_extent_yrs_ch_CON)
```

## Appendix 2

```{r}
app2 <- ggplot(RF_data %>%
         mutate(hab = ifelse(eunis_cambio == F, eunis1, eunis_first)) %>%
         mutate(cat = factor(ifelse(eunis_cambio == FALSE, "Constant", "Change"),
                             levels = c("Constant", "Change"))), 
       aes(x = hab, y = delta_focal_pct, fill = hab)) +
  geom_point(aes(color = hab, alpha = cat),
             position = position_jitter(width = 0.25), size = 1) +
  scale_alpha_manual(
    values = c("Constant" = 0.1, "Change" = 0.5), guide = "none") +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2,
               linewidth = 0.3, color = "black") +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 2,
               color = "black", fill = "black") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  facet_grid(rows = vars(cat), cols = vars(biogeo)) +
  labs(x = "EUNIS I Habitat type", y = "Change in habitat extent (%)") +
  theme_bw() +
  theme(legend.position = "none") 
app2
ggsave(app2, filename = "output/figures/app2.jpeg", 
       width = 30, height = 24, units = "cm", dpi = 300)
```

# Change in functional metrics

## Correlations among functional metrics

```{r}
RF_data %>%
  dplyr::select(matches("^(ndvi|savi|ndmi|ndwi)")) %>%
  cor(use = "pairwise.complete.obs") %>%
  corrplot(method = "color", type = "upper", addCoef.col = "black",
           number.cex = 1, tl.col = "black")
```

## Plots with constant habitat

### Change in functional metrics by habitat type

```{r}
mod_ndvi_eunis1_noch <- lm(ndvi_resurvey ~ eunis1, 
                           data = RF_data %>% filter(eunis_cambio == F))
mod_savi_eunis1_noch <- lm(savi_resurvey ~ eunis1, 
                           data = RF_data %>% filter(eunis_cambio == F))
mod_ndmi_eunis1_noch <- lm(ndmi_resurvey ~ eunis1, 
                           data = RF_data %>% filter(eunis_cambio == F))
mod_ndwi_eunis1_noch <- lm(ndwi_resurvey ~ eunis1, 
                           data = RF_data %>% filter(eunis_cambio == F))
Anova(mod_ndvi_eunis1_noch)
Anova(mod_savi_eunis1_noch)
Anova(mod_ndmi_eunis1_noch)
Anova(mod_ndwi_eunis1_noch)
```

```{r}
emmeans(mod_ndvi_eunis1_noch, pairwise ~ eunis1)
emmeans(mod_savi_eunis1_noch, pairwise ~ eunis1)
emmeans(mod_ndmi_eunis1_noch, pairwise ~ eunis1)
emmeans(mod_ndwi_eunis1_noch, pairwise ~ eunis1)
```

Proportion of variance explained:

```{r}
summary(mod_ndvi_eunis1_noch)$r.squared
summary(mod_savi_eunis1_noch)$r.squared
summary(mod_ndmi_eunis1_noch)$r.squared
summary(mod_ndwi_eunis1_noch)$r.squared
```

### Figure 2

```{r}
fig2 <- ggplot(RF_data %>% filter(eunis_cambio == F) %>%
                  pivot_longer(cols = c(ndvi_resurvey, savi_resurvey,
                                        ndmi_resurvey, ndwi_resurvey), 
                               names_to = "index", values_to = "value") %>%
                  mutate(index = factor(index, 
                                        levels = c("ndvi_resurvey", 
                                                   "savi_resurvey",
                                                   "ndmi_resurvey", 
                                                   "ndwi_resurvey"),
                                        labels = c("NDVI", "SAVI", "NDMI", 
                                                   "NDWI"))), 
                aes(x = eunis1, y = value, fill = eunis1)) +
  geom_point(aes(color = eunis1),
             position = position_jitter(width = 0.25), size = 0.5, alpha = 0.1) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2,
                linewidth = 0.3, color = "black") +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 2,
                color = "black", fill = "black") +
  facet_wrap(~ index, ncol = 2, scales = "free") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "EUNIS I Habitat type", y = "Sen's slope") +
  theme_bw() +
  theme(legend.position = "none") 
fig2
ggsave(fig2, filename = "output/figures/fig2.jpeg", 
       width = 24, height = 24, units = "cm", dpi = 300)
```

## Level II

```{r}
# Prepare data
plot_data_eunis2 <- RF_data %>%
  filter(eunis_cambio == FALSE) %>%
  filter(eunis2 != "Change") %>%
  pivot_longer(cols = c(ndvi_resurvey, savi_resurvey, ndmi_resurvey, 
                        ndwi_resurvey),
               names_to = "index", values_to = "value") %>%
  mutate(index = factor(index,
                        levels = c("ndvi_resurvey", "savi_resurvey",
                                   "ndmi_resurvey", "ndwi_resurvey"),
                        labels = c("NDVI", "SAVI", "NDMI", "NDWI")))

# Sample sizes per EUNIS II class
eunis2_counts <- plot_data_eunis2 %>%
  distinct(plot_id, eunis2) %>%   # avoids counting same plot several times
  count(eunis2)

# Add labels containing habitat name + n
plot_data_eunis2 <- plot_data_eunis2 %>%
  left_join(eunis2_counts, by = "eunis2") %>%
  mutate(eunis2_label = paste0(eunis2, "\n(n=", n, ")"))
```

```{r}
ggplot(plot_data_eunis2 %>%
         # Keep eunis2 with > 10 observations
         filter(n > 10) %>%
         filter(index == "NDVI"), 
       aes(x = eunis2_label, y = value, fill = eunis1)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_point(aes(color = eunis1), position = position_jitter(width = 0.25),
             size = 0.5, alpha = 0.3) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2,
               linewidth = 0.3, color = "black") +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 2, 
               color = "black", fill = "black") +
  facet_wrap(~index, ncol = 1, scales = "free") +
  labs(x = "EUNIS II habitat type", y = "Sen's slope") +
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(size = 8, lineheight = 0.9))
```

```{r}
ggplot(plot_data_eunis2 %>%
         # Keep eunis2 with > 10 observations
         filter(n > 10) %>%
         filter(index == "SAVI"), 
       aes(x = eunis2_label, y = value, fill = eunis1)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_point(aes(color = eunis1), position = position_jitter(width = 0.25),
             size = 0.5, alpha = 0.3) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2,
               linewidth = 0.3, color = "black") +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 2, 
               color = "black", fill = "black") +
  facet_wrap(~index, ncol = 1, scales = "free") +
  labs(x = "EUNIS II habitat type", y = "Sen's slope") +
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(size = 8, lineheight = 0.9))
```

```{r}
ggplot(plot_data_eunis2 %>%
         # Keep eunis2 with > 10 observations
         filter(n > 10) %>%
         filter(index == "NDMI"), 
       aes(x = eunis2_label, y = value, fill = eunis1)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_point(aes(color = eunis1), position = position_jitter(width = 0.25),
             size = 0.5, alpha = 0.3) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2,
               linewidth = 0.3, color = "black") +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 2, 
               color = "black", fill = "black") +
  facet_wrap(~index, ncol = 1, scales = "free") +
  labs(x = "EUNIS II habitat type", y = "Sen's slope") +
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(size = 8, lineheight = 0.9))
```

```{r}
ggplot(plot_data_eunis2 %>%
         # Keep eunis2 with > 10 observations
         filter(n > 10) %>%
         filter(index == "NDWI"), 
       aes(x = eunis2_label, y = value, fill = eunis1)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_point(aes(color = eunis1), position = position_jitter(width = 0.25),
             size = 0.5, alpha = 0.3) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2,
               linewidth = 0.3, color = "black") +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 2, 
               color = "black", fill = "black") +
  facet_wrap(~index, ncol = 1, scales = "free") +
  labs(x = "EUNIS II habitat type", y = "Sen's slope") +
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(size = 8, lineheight = 0.9))
```

### SAVI & NDWI in R

```{r}
ggplot(plot_data_eunis2 %>%
         # Keep eunis2 with > 10 observations
         filter(n > 10) %>%
         filter(index %in% c("SAVI", "NDWI") & eunis1 == "R"), 
       aes(x = eunis2_label, y = value)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_point(position = position_jitter(width = 0.25),
             size = 1, alpha = 0.3, color = "#7CAE00") +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2,
               linewidth = 0.3, color = "black") +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 2, 
               color = "black", fill = "black") +
  facet_wrap(~index, ncol = 2, scales = "free") +
  labs(x = "EUNIS II habitat type", y = "Sen's slope") +
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(size = 8, lineheight = 0.9))
```

### NDMI & NDWI in R

```{r}
ggplot(plot_data_eunis2 %>%
         # Keep eunis2 with > 10 observations
         filter(n > 10) %>%
         filter(index %in% c("NDMI", "NDWI") & eunis1 == "Q"), 
       aes(x = eunis2_label, y = value)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_point(position = position_jitter(width = 0.25),
             size = 1, alpha = 0.4, color = "#F8766D") +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2,
               linewidth = 0.3, color = "black") +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 2, 
               color = "black", fill = "black") +
  facet_wrap(~index, ncol = 2, scales = "free") +
  labs(x = "EUNIS II habitat type", y = "Sen's slope") +
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(size = 8, lineheight = 0.9))
```

### Test with R

```{r}
mod_ndvi_eunis2R_noch <- lm(ndvi_resurvey ~ eunis2, 
                           data = RF_data %>% 
                             filter(eunis_cambio == F & eunis1 == "R" &
                                      eunis2 != "Change"))
mod_savi_eunis2R_noch <- lm(savi_resurvey ~ eunis2, 
                           data = RF_data %>% 
                             filter(eunis_cambio == F & eunis1 == "R" &
                                      eunis2 != "Change"))
mod_ndmi_eunis2R_noch <- lm(ndmi_resurvey ~ eunis2, 
                           data = RF_data %>% 
                             filter(eunis_cambio == F & eunis1 == "R" &
                                      eunis2 != "Change"))
mod_ndwi_eunis2R_noch <- lm(ndwi_resurvey ~ eunis2, 
                           data = RF_data %>% 
                             filter(eunis_cambio == F & eunis1 == "R" &
                                      eunis2 != "Change"))
Anova(mod_ndvi_eunis2R_noch)
Anova(mod_savi_eunis2R_noch)
Anova(mod_ndmi_eunis2R_noch)
Anova(mod_ndwi_eunis2R_noch)
```

```{r}
emmeans(mod_ndvi_eunis2R_noch, pairwise ~ eunis2)
emmeans(mod_savi_eunis2R_noch, pairwise ~ eunis2)
emmeans(mod_ndmi_eunis2R_noch, pairwise ~ eunis2)
emmeans(mod_ndwi_eunis2R_noch, pairwise ~ eunis2)
```

### Test with Q

```{r}
mod_ndvi_eunis2Q_noch <- lm(ndvi_resurvey ~ eunis2, 
                           data = RF_data %>% 
                             filter(eunis_cambio == F & eunis1 == "Q" &
                                      eunis2 != "Change"))
mod_savi_eunis2Q_noch <- lm(savi_resurvey ~ eunis2, 
                           data = RF_data %>% 
                             filter(eunis_cambio == F & eunis1 == "Q" &
                                      eunis2 != "Change"))
mod_ndmi_eunis2Q_noch <- lm(ndmi_resurvey ~ eunis2, 
                           data = RF_data %>% 
                           filter(eunis_cambio == F & eunis1 == "Q" &
                                      eunis2 != "Change"))
mod_ndwi_eunis2Q_noch <- lm(ndwi_resurvey ~ eunis2, 
                           data = RF_data %>% 
                             filter(eunis_cambio == F & eunis1 == "Q" &
                                      eunis2 != "Change"))
Anova(mod_ndvi_eunis2Q_noch)
Anova(mod_savi_eunis2Q_noch)
Anova(mod_ndmi_eunis2Q_noch)
Anova(mod_ndwi_eunis2Q_noch)
```

```{r}
emmeans(mod_ndmi_eunis2Q_noch, pairwise ~ eunis2)
emmeans(mod_ndwi_eunis2Q_noch, pairwise ~ eunis2)
```

## Plots with habitat change

### Change in functional metrics by habitat type

```{r}
mod_ndvi_eunis1_ch <- lm(ndvi_resurvey ~ eunis_first, 
                         data = RF_data %>% filter(eunis_cambio == T))
mod_savi_eunis1_ch <- lm(savi_resurvey ~ eunis_first, 
                         data = RF_data %>% filter(eunis_cambio == T))
mod_ndmi_eunis1_ch <- lm(ndmi_resurvey ~ eunis_first, 
                         data = RF_data %>% filter(eunis_cambio == T))
mod_ndwi_eunis1_ch <- lm(ndwi_resurvey ~ eunis_first, 
                         data = RF_data %>% filter(eunis_cambio == T))
Anova(mod_ndvi_eunis1_ch)
Anova(mod_savi_eunis1_ch)
Anova(mod_ndmi_eunis1_ch)
Anova(mod_ndwi_eunis1_ch)
```

```{r}
emmeans(mod_ndvi_eunis1_ch, pairwise ~ eunis_first)
emmeans(mod_savi_eunis1_ch, pairwise ~ eunis_first)
emmeans(mod_ndmi_eunis1_ch, pairwise ~ eunis_first)
emmeans(mod_ndwi_eunis1_ch, pairwise ~ eunis_first)
```

Proportion of variance explained:

```{r}
summary(mod_ndvi_eunis1_ch)$r.squared
summary(mod_savi_eunis1_ch)$r.squared
summary(mod_ndmi_eunis1_ch)$r.squared
summary(mod_ndwi_eunis1_ch)$r.squared
```

#### Appendix 3

```{r}
app3 <- ggplot(RF_data %>% filter(eunis_cambio == T) %>%
                  pivot_longer(cols = c(ndvi_resurvey, savi_resurvey,
                                        ndmi_resurvey, ndwi_resurvey), 
                               names_to = "index", values_to = "value") %>%
                  mutate(index = factor(index, 
                                        levels = c("ndvi_resurvey", 
                                                   "savi_resurvey",
                                                   "ndmi_resurvey", 
                                                   "ndwi_resurvey"),
                                        labels = c("NDVI", "SAVI", "NDMI", 
                                                   "NDWI"))), 
                aes(x = eunis_first, y = value, fill = eunis_first)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_point(aes(color = eunis_first),
             position = position_jitter(width = 0.25), size = 2, alpha = 0.2) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2,
               linewidth = 0.3, color = "black") +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 2,
                color = "black", fill = "black") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  facet_wrap(~ index, nrow = 2, scales = "free") +
  labs(x = "Original EUNIS I Habitat type", y = "Sen's slope") +
  theme_bw() +
  theme(legend.position = "none") 
app3
ggsave(app3, filename = "output/figures/app3.jpeg", 
       width = 24, height = 24, units = "cm", dpi = 300)
```

### Change in functional metrics by transition type

```{r}
ggplot(RF_data %>% filter(eunis_cambio == TRUE) %>%
         left_join(RF_data %>% filter(eunis_cambio == TRUE) %>%
                     count(transition),
                   by = "transition") %>%
         mutate(transition_label = paste0(transition, "\n(n = ", n, ")")) %>%
         pivot_longer(cols = c(ndvi_resurvey, savi_resurvey, ndmi_resurvey, 
                               ndwi_resurvey),
                      names_to = "index", values_to = "value") %>%
         mutate(index = factor(index, 
                               levels = c("ndvi_resurvey", "savi_resurvey",
                                          "ndmi_resurvey", "ndwi_resurvey"),
                               labels = c("NDVI", "SAVI", "NDMI", "NDWI"))), 
       aes(x = transition_label, y = value, fill = eunis_first)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_point(aes(color = eunis_first), position = position_jitter(width = 0.25),
             size = 2, alpha = 0.2) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2,
               linewidth = 0.3, color = "black") +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 2, 
               color = "black", fill = "black") +
  facet_wrap(~index, nrow = 2, scales = "free") +
  labs(x = "Transition", y = "Sen's slope") +
  theme_bw() + theme(axis.text.x = element_text(size = 8, lineheight = 0.9)) +
  theme(legend.position = "none")
```


# Change in functional metrics by bioregion

## Plots with constant habitat

```{r}
table((RF_data %>% filter(eunis_cambio == F))$eunis1, 
      (RF_data %>% filter(eunis_cambio == F))$biogeo)
```

### Change in functional metrics by habitat type 

#### NDVI

```{r}
mod_ndvi_eunis1_noch_ALP <- lm(ndvi_resurvey ~ eunis1,
                                 data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                   filter(biogeo == "ALP"))
mod_ndvi_eunis1_noch_ATL <- lm(ndvi_resurvey ~ eunis1,
                                 data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                   filter(biogeo == "ATL"))
mod_ndvi_eunis1_noch_CON <- lm(ndvi_resurvey ~ eunis1,
                                 data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                   filter(biogeo == "CON"))
# Not possible in MED because all are R
mod_ndvi_eunis1_noch_PAN <- lm(ndvi_resurvey ~ eunis1,
                                 data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                   filter(biogeo == "PAN"))
Anova(mod_ndvi_eunis1_noch_ALP)
Anova(mod_ndvi_eunis1_noch_ATL)
Anova(mod_ndvi_eunis1_noch_CON)
Anova(mod_ndvi_eunis1_noch_PAN)
```

```{r}
emmeans(mod_ndvi_eunis1_noch_ATL, pairwise ~ eunis1)
emmeans(mod_ndvi_eunis1_noch_CON, pairwise ~ eunis1)
```

Proportion of variance explained:

```{r}
summary(mod_ndvi_eunis1_noch_ALP)$r.squared
summary(mod_ndvi_eunis1_noch_ATL)$r.squared
summary(mod_ndvi_eunis1_noch_CON)$r.squared
summary(mod_ndvi_eunis1_noch_PAN)$r.squared
```

#### SAVI

```{r}
mod_savi_eunis1_noch_ALP <- lm(savi_resurvey ~ eunis1,
                                 data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                   filter(biogeo == "ALP"))
mod_savi_eunis1_noch_ATL <- lm(savi_resurvey ~ eunis1,
                                 data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                   filter(biogeo == "ATL"))
mod_savi_eunis1_noch_CON <- lm(savi_resurvey ~ eunis1,
                                 data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                   filter(biogeo == "CON"))
# Not possible in MED because all are R
mod_savi_eunis1_noch_PAN <- lm(savi_resurvey ~ eunis1,
                                 data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                   filter(biogeo == "PAN"))
Anova(mod_savi_eunis1_noch_ALP)
Anova(mod_savi_eunis1_noch_ATL)
Anova(mod_savi_eunis1_noch_CON)
Anova(mod_savi_eunis1_noch_PAN)
```

```{r}
emmeans(mod_savi_eunis1_noch_ATL, pairwise ~ eunis1)
emmeans(mod_savi_eunis1_noch_CON, pairwise ~ eunis1)
emmeans(mod_savi_eunis1_noch_PAN, pairwise ~ eunis1)
```

Proportion of variance explained:

```{r}
summary(mod_savi_eunis1_noch_ALP)$r.squared
summary(mod_savi_eunis1_noch_ATL)$r.squared
summary(mod_savi_eunis1_noch_CON)$r.squared
summary(mod_savi_eunis1_noch_PAN)$r.squared
```

#### NDMI

```{r}
mod_ndmi_eunis1_noch_ALP <- lm(ndmi_resurvey ~ eunis1,
                                 data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                   filter(biogeo == "ALP"))
mod_ndmi_eunis1_noch_ATL <- lm(ndmi_resurvey ~ eunis1,
                                 data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                   filter(biogeo == "ATL"))
mod_ndmi_eunis1_noch_CON <- lm(ndmi_resurvey ~ eunis1,
                                 data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                   filter(biogeo == "CON"))
# Not possible in MED because all are R
mod_ndmi_eunis1_noch_PAN <- lm(ndmi_resurvey ~ eunis1,
                                 data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                   filter(biogeo == "PAN"))
Anova(mod_ndmi_eunis1_noch_ALP)
Anova(mod_ndmi_eunis1_noch_ATL)
Anova(mod_ndmi_eunis1_noch_CON)
Anova(mod_ndmi_eunis1_noch_PAN)
```

```{r}
emmeans(mod_ndmi_eunis1_noch_ATL, pairwise ~ eunis1)
emmeans(mod_ndmi_eunis1_noch_CON, pairwise ~ eunis1)
```

Proportion of variance explained:

```{r}
summary(mod_ndmi_eunis1_noch_ALP)$r.squared
summary(mod_ndmi_eunis1_noch_ATL)$r.squared
summary(mod_ndmi_eunis1_noch_CON)$r.squared
summary(mod_ndmi_eunis1_noch_PAN)$r.squared
```

#### NDWI

```{r}
mod_ndwi_eunis1_noch_ALP <- lm(ndwi_resurvey ~ eunis1,
                                 data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                   filter(biogeo == "ALP"))
mod_ndwi_eunis1_noch_ATL <- lm(ndwi_resurvey ~ eunis1,
                                 data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                   filter(biogeo == "ATL"))
mod_ndwi_eunis1_noch_CON <- lm(ndwi_resurvey ~ eunis1,
                                 data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                   filter(biogeo == "CON"))
# Not possible in MED because all are R
mod_ndwi_eunis1_noch_PAN <- lm(ndwi_resurvey ~ eunis1,
                                 data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                   filter(biogeo == "PAN"))
Anova(mod_ndwi_eunis1_noch_ALP)
Anova(mod_ndwi_eunis1_noch_ATL)
Anova(mod_ndwi_eunis1_noch_CON)
Anova(mod_ndwi_eunis1_noch_PAN)
```

```{r}
emmeans(mod_ndwi_eunis1_noch_ATL, pairwise ~ eunis1)
emmeans(mod_ndwi_eunis1_noch_CON, pairwise ~ eunis1)
emmeans(mod_ndwi_eunis1_noch_PAN, pairwise ~ eunis1)
```

Proportion of variance explained:

```{r}
summary(mod_ndwi_eunis1_noch_ALP)$r.squared
summary(mod_ndwi_eunis1_noch_ATL)$r.squared
summary(mod_ndwi_eunis1_noch_CON)$r.squared
summary(mod_ndwi_eunis1_noch_PAN)$r.squared
```

### Change in functional metrics by number of years between first and last survey

#### NDVI

```{r}
mod_ndvi_yrs_noch_ALP <- lm(abs(ndvi_resurvey) ~ years_between.x,
                                     data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                       filter(biogeo == "ALP"))
mod_ndvi_yrs_noch_ATL <- lm(abs(ndvi_resurvey) ~ years_between.x,
                                     data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                       filter(biogeo == "ATL"))
mod_ndvi_yrs_noch_CON <- lm(abs(ndvi_resurvey) ~ years_between.x,
                                     data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                       filter(biogeo == "CON"))
mod_ndvi_yrs_noch_MED <- lm(abs(ndvi_resurvey) ~ years_between.x,
                                     data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                       filter(biogeo == "MED"))
mod_ndvi_yrs_noch_PAN <- lm(abs(ndvi_resurvey) ~ years_between.x,
                                     data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                       filter(biogeo == "PAN"))
Anova(mod_ndvi_yrs_noch_ALP)
Anova(mod_ndvi_yrs_noch_ATL)
Anova(mod_ndvi_yrs_noch_CON)
Anova(mod_ndvi_yrs_noch_MED)
Anova(mod_ndvi_yrs_noch_PAN)
summary(mod_ndvi_yrs_noch_ALP)
summary(mod_ndvi_yrs_noch_ATL)
summary(mod_ndvi_yrs_noch_CON)
summary(mod_ndvi_yrs_noch_MED)
summary(mod_ndvi_yrs_noch_PAN)
```

#### SAVI

```{r}
mod_savi_yrs_noch_ALP <- lm(abs(savi_resurvey) ~ years_between.x,
                            data = RF_data %>% 
                              filter(eunis_cambio == F) %>%
                              filter(biogeo == "ALP"))
mod_savi_yrs_noch_ATL <- lm(abs(savi_resurvey) ~ years_between.x,
                            data = RF_data %>% 
                              filter(eunis_cambio == F) %>%
                              filter(biogeo == "ATL"))
mod_savi_yrs_noch_CON <- lm(abs(savi_resurvey) ~ years_between.x,
                                     data = RF_data %>% 
                              filter(eunis_cambio == F) %>%
                              filter(biogeo == "CON"))
mod_savi_yrs_noch_MED <- lm(abs(savi_resurvey) ~ years_between.x,
                                     data = RF_data %>% 
                              filter(eunis_cambio == F) %>%
                              filter(biogeo == "MED"))
mod_savi_yrs_noch_PAN <- lm(abs(savi_resurvey) ~ years_between.x,
                                     data = RF_data %>% 
                              filter(eunis_cambio == F) %>%
                              filter(biogeo == "PAN"))
Anova(mod_savi_yrs_noch_ALP)
Anova(mod_savi_yrs_noch_ATL)
Anova(mod_savi_yrs_noch_CON)
Anova(mod_savi_yrs_noch_MED)
Anova(mod_savi_yrs_noch_PAN)
summary(mod_savi_yrs_noch_ALP)
summary(mod_savi_yrs_noch_ATL)
summary(mod_savi_yrs_noch_CON)
summary(mod_savi_yrs_noch_MED)
summary(mod_savi_yrs_noch_PAN)
```

#### NDMI

```{r}
mod_ndmi_yrs_noch_ALP <- lm(abs(ndmi_resurvey) ~ years_between.x,
                                     data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                       filter(biogeo == "ALP"))
mod_ndmi_yrs_noch_ATL <- lm(abs(ndmi_resurvey) ~ years_between.x,
                                     data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                       filter(biogeo == "ATL"))
mod_ndmi_yrs_noch_CON <- lm(abs(ndmi_resurvey) ~ years_between.x,
                                     data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                       filter(biogeo == "CON"))
mod_ndmi_yrs_noch_MED <- lm(abs(ndmi_resurvey) ~ years_between.x,
                                     data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                       filter(biogeo == "MED"))
mod_ndmi_yrs_noch_PAN <- lm(abs(ndmi_resurvey) ~ years_between.x,
                                     data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                       filter(biogeo == "PAN"))
Anova(mod_ndmi_yrs_noch_ALP)
Anova(mod_ndmi_yrs_noch_ATL)
Anova(mod_ndmi_yrs_noch_CON)
Anova(mod_ndmi_yrs_noch_MED)
Anova(mod_ndmi_yrs_noch_PAN)
summary(mod_ndmi_yrs_noch_ALP)
summary(mod_ndmi_yrs_noch_ATL)
summary(mod_ndmi_yrs_noch_CON)
summary(mod_ndmi_yrs_noch_MED)
summary(mod_ndmi_yrs_noch_PAN)
```

#### NDWI

```{r}
mod_ndwi_yrs_noch_ALP <- lm(abs(ndwi_resurvey) ~ years_between.x,
                                     data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                       filter(biogeo == "ALP"))
mod_ndwi_yrs_noch_ATL <- lm(abs(ndwi_resurvey) ~ years_between.x,
                                     data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                       filter(biogeo == "ATL"))
mod_ndwi_yrs_noch_CON <- lm(abs(ndwi_resurvey) ~ years_between.x,
                                     data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                       filter(biogeo == "CON"))
mod_ndwi_yrs_noch_MED <- lm(abs(ndwi_resurvey) ~ years_between.x,
                                     data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                       filter(biogeo == "MED"))
mod_ndwi_yrs_noch_PAN <- lm(abs(ndwi_resurvey) ~ years_between.x,
                                     data = RF_data %>% 
                                 filter(eunis_cambio == F) %>%
                                       filter(biogeo == "PAN"))
Anova(mod_ndwi_yrs_noch_ALP)
Anova(mod_ndwi_yrs_noch_ATL)
Anova(mod_ndwi_yrs_noch_CON)
Anova(mod_ndwi_yrs_noch_MED)
Anova(mod_ndwi_yrs_noch_PAN)
summary(mod_ndwi_yrs_noch_ALP)
summary(mod_ndwi_yrs_noch_ATL)
summary(mod_ndwi_yrs_noch_CON)
summary(mod_ndwi_yrs_noch_MED)
summary(mod_ndwi_yrs_noch_PAN)
```

### Appendix 4

```{r}
app4 <- ggplot(RF_data %>% filter(eunis_cambio == F) %>%
                 pivot_longer(cols = c(ndvi_resurvey, savi_resurvey,
                                        ndmi_resurvey, ndwi_resurvey), 
                               names_to = "index", values_to = "value") %>%
                  mutate(index = factor(index, 
                                        levels = c("ndvi_resurvey", 
                                                   "savi_resurvey",
                                                   "ndmi_resurvey", 
                                                   "ndwi_resurvey"),
                                        labels = c("NDVI", "SAVI", "NDMI", 
                                                   "NDWI"))), 
                aes(x = eunis1, y = value, fill = eunis1)) +
  geom_point(aes(color = eunis1),
             position = position_jitter(width = 0.25), size = 1, alpha = 0.1) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2,
                linewidth = 0.3, color = "black") +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 2,
                color = "black", fill = "black") +
  facet_grid(rows = vars(index), cols = vars(biogeo), scales = "free") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "EUNIS I Habitat type", y = "Sen's slope") +
  theme_bw() +
  theme(legend.position = "none") 
app4
ggsave(app4, filename = "output/figures/app4.jpeg", 
       width = 36, height = 36, units = "cm", dpi = 300)
```

## Plots with habitat change

```{r}
table((RF_data %>% filter(eunis_cambio == T))$eunis_first,
      (RF_data %>% filter(eunis_cambio == T))$biogeo)
```

### Change in functional metrics by habitat type 

#### NDVI

```{r}
mod_ndvi_eunis_first_ch_ALP <- lm(ndvi_resurvey ~ eunis_first,
                                 data = RF_data %>% 
                                   filter(eunis_cambio == T) %>%
                                   filter(biogeo == "ALP"))
mod_ndvi_eunis_first_ch_ATL <- lm(ndvi_resurvey ~ eunis_first,
                                 data = RF_data %>% 
                                   filter(eunis_cambio == T) %>%
                                   filter(biogeo == "ATL"))
mod_ndvi_eunis_first_ch_CON <- lm(ndvi_resurvey ~ eunis_first,
                                 data = RF_data %>% 
                                   filter(eunis_cambio == T) %>%
                                   filter(biogeo == "CON"))
Anova(mod_ndvi_eunis_first_ch_ALP)
Anova(mod_ndvi_eunis_first_ch_ATL)
Anova(mod_ndvi_eunis_first_ch_CON)
```

```{r}
emmeans(mod_ndvi_eunis_first_ch_CON, pairwise ~ eunis_first)
```


Proportion of variance explained:

```{r}
summary(mod_ndvi_eunis_first_ch_ALP)$r.squared
summary(mod_ndvi_eunis_first_ch_ATL)$r.squared
summary(mod_ndvi_eunis_first_ch_CON)$r.squared
```

#### SAVI

```{r}
mod_savi_eunis_first_ch_ALP <- lm(savi_resurvey ~ eunis_first,
                                 data = RF_data %>% 
                                   filter(eunis_cambio == T) %>%
                                   filter(biogeo == "ALP"))
mod_savi_eunis_first_ch_ATL <- lm(savi_resurvey ~ eunis_first,
                                 data = RF_data %>% 
                                   filter(eunis_cambio == T) %>%
                                   filter(biogeo == "ATL"))
mod_savi_eunis_first_ch_CON <- lm(savi_resurvey ~ eunis_first,
                                 data = RF_data %>% 
                                   filter(eunis_cambio == T) %>%
                                   filter(biogeo == "CON"))
Anova(mod_savi_eunis_first_ch_ALP)
Anova(mod_savi_eunis_first_ch_ATL)
Anova(mod_savi_eunis_first_ch_CON)
```

```{r}
emmeans(mod_savi_eunis_first_ch_CON, pairwise ~ eunis_first)
```


Proportion of variance explained:

```{r}
summary(mod_savi_eunis_first_ch_ALP)$r.squared
summary(mod_savi_eunis_first_ch_ATL)$r.squared
summary(mod_savi_eunis_first_ch_CON)$r.squared
```

#### NDMI

```{r}
mod_ndmi_eunis_first_ch_ALP <- lm(ndmi_resurvey ~ eunis_first,
                                 data = RF_data %>% 
                                   filter(eunis_cambio == T) %>%
                                   filter(biogeo == "ALP"))
mod_ndmi_eunis_first_ch_ATL <- lm(ndmi_resurvey ~ eunis_first,
                                 data = RF_data %>% 
                                   filter(eunis_cambio == T) %>%
                                   filter(biogeo == "ATL"))
mod_ndmi_eunis_first_ch_CON <- lm(ndmi_resurvey ~ eunis_first,
                                 data = RF_data %>% 
                                   filter(eunis_cambio == T) %>%
                                   filter(biogeo == "CON"))
Anova(mod_ndmi_eunis_first_ch_ALP)
Anova(mod_ndmi_eunis_first_ch_ATL)
Anova(mod_ndmi_eunis_first_ch_CON)
```

```{r}
emmeans(mod_ndmi_eunis_first_ch_CON, pairwise ~ eunis_first)
```


Proportion of variance explained:

```{r}
summary(mod_ndmi_eunis_first_ch_ALP)$r.squared
summary(mod_ndmi_eunis_first_ch_ATL)$r.squared
summary(mod_ndmi_eunis_first_ch_CON)$r.squared
```

#### NDWI

```{r}
mod_ndwi_eunis_first_ch_ALP <- lm(ndwi_resurvey ~ eunis_first,
                                 data = RF_data %>% 
                                   filter(eunis_cambio == T) %>%
                                   filter(biogeo == "ALP"))
mod_ndwi_eunis_first_ch_ATL <- lm(ndwi_resurvey ~ eunis_first,
                                 data = RF_data %>% 
                                   filter(eunis_cambio == T) %>%
                                   filter(biogeo == "ATL"))
mod_ndwi_eunis_first_ch_CON <- lm(ndwi_resurvey ~ eunis_first,
                                 data = RF_data %>% 
                                   filter(eunis_cambio == T) %>%
                                   filter(biogeo == "CON"))
Anova(mod_ndwi_eunis_first_ch_ALP)
Anova(mod_ndwi_eunis_first_ch_ATL)
Anova(mod_ndwi_eunis_first_ch_CON)
```

```{r}
emmeans(mod_ndwi_eunis_first_ch_CON, pairwise ~ eunis_first)
```

Proportion of variance explained:

```{r}
summary(mod_ndwi_eunis_first_ch_ALP)$r.squared
summary(mod_ndwi_eunis_first_ch_ATL)$r.squared
summary(mod_ndwi_eunis_first_ch_CON)$r.squared
```

# Session info

```{r}
sessionInfo()
```

