---
title: "MOTIVATE mapping - Analysing biodiversity - habitat exent (RF) relationships"
author: "Alicia Valdés"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: 4
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  word_document:
    toc: yes
    toc_depth: '4'
editor_options:
  chunk_output_type: inline
---

This notebook analyses relationships between biodiversity metrics and habitat extent metrics calculated from RF.

# Load libraries

```{r}
library(dplyr)
library(tidyr)
library(readr)
library(here)
library(lubridate)
library(stringr)
library(purrr)
library(broom)
library(MASS)
library(ggplot2)
library(corrplot)
library(performance)
library(robustbase)
```

# Load data on habitat extent from random forest models

```{r}
RF_data <- readRDS(here("data", "clean", "RF_data.rds"))
```

# Load data from Nat. Comms. paper

From https://idata.idiv.de/ddm/Data/ShowData/3611

A pan-Europe dataset of estimated linear trends in multiple dimension of local plant diversity in 57,390 vegetation-plot time series.

```{r}
biodiv_perc <- read_tsv(here("data", "all_lm_trends_percentage_change_scale.txt"))
biodiv_perc
```

# Load data on CWMs

Data on percentage annual changes in CWMs per time_seriesID                                                    :

```{r}
CWMs <- read_rds(here("data", "CWMs", "all_rate_lm_tidy_fits_CWMs_for_Alicia.rds"))
```

```{r}
CWMs %>% distinct(metric)
```

CWM_PlantHeight.veg: CWM for plant height
CWM_RootingDepth: CWM for rooting depth			
CWM_LeafCN.ratio: CWM for leaf C:N ratio
CWM_LeafArea.leaf.undef: CWM for leaf area
CWM_SLA.undef: CWM for specific leaf area (SLA)

Rename to easier names:

```{r}
CWMs <- CWMs %>%
  mutate(metric = case_when(
    metric == "CWM_PlantHeight.veg" ~ "CWM_PlantHeight",
    metric == "CWM_RootingDepth" ~ "CWM_RootingDepth",
    metric == "CWM_LeafCN.ratio" ~ "CWM_LeafCNratio",
    metric == "CWM_LeafArea.leaf.undef" ~ "CWM_LeafArea",
    metric == "CWM_SLA.undef" ~ "CWM_SLA",
    TRUE ~ metric
  ))
```

# Load original data

```{r}
data_V2025.2 <- read_tsv(
  "C:/Data/MOTIVATE/MOTIVATE_data/Data/EVA and ReSurvey Data/2025 data - V2025.2/Core data/200_Motivate20250415_notJUICE_header_with_precise_coordinates.csv") %>%
  mutate(date = dmy(`Date of recording`), year = as.integer(year(date)))
```

# Data wrangling

## Join CMWs data to biodiversity data

```{r}
biodiv_perc <- bind_rows(biodiv_perc, 
                         CWMs %>%
                           dplyr::select(metric, time_seriesID, slope,
                                         initial_div_value, EUNIS_start, 
                                         EUNIS_end, year_start, year_end, n, 
                                         Trajectory, plot_size))
```

## Add columns for EUNIS1 (start and end) to biodiversity data

```{r}
biodiv_perc <- biodiv_perc %>%
  mutate(EUNIS1_start = str_sub(EUNIS_start, 1, 1),
         EUNIS1_end = str_sub(EUNIS_end, 1, 1))
```

## Fix plotIDs in RF_data

Gonzalo used RS_CODE instead of Resurvey project as first part of plot_id.

Retrieve list of correspondences of RS_CODE and Resurvey project from original data:

```{r}
correspondence_list <- data_V2025.2 %>%
  distinct(RS_CODE, `ReSurvey project`)
```

```{r}
RF_data <- RF_data %>%
  # Add spaces before and after "-"
  mutate(plotID = str_replace_all(plot_id, "-", " - ")) %>%
  # Get RS_CODE: first part of plot_id, before first "-"
  mutate(RS_CODE = sub("-.*", "", plot_id)) %>%
  # Join to get Resurvey project
  left_join(correspondence_list) %>%
  # Construct new plotID
  mutate(plotID = str_replace(plotID, paste0("^", RS_CODE), 
                              `ReSurvey project`)) %>%
  # Remove unneeded columns
  dplyr::select(-plot_id, -RS_CODE, -`ReSurvey project`) %>%
  # Modify some plotIDs
  mutate(
    plotID = str_replace(
      plotID,
      "^AOPK_HabitatMonitoring - PolickeSteny - Signal -",
      "AOPK_HabitatMonitoring - PolickeSteny-Signal -"
      ),
    plotID = str_replace(
      plotID,
      "^Ecohydrology of Floodplain Meadows - ([0-9]+) - \\1\\s+-\\s+",
      "Ecohydrology of Floodplain Meadows - \\1 - \\1-"
      ),
    plotID = str_replace(
      plotID,
      "^PSET Puy Long - PuyLong - PSET 0 - 1",
      "PSET Puy Long - PuyLong - PSET 0-1"
      ),
    plotID = str_replace(
      plotID,
      "^PSET Puy Long - PuyLong - PSET ([0-9]+) - ([0-9]+)",
      "PSET Puy Long - PuyLong - PSET \\1-\\2"
      ),
    plotID = str_replace(
      plotID,
      "^Role of AES schemes on the grasslands biodiversity - ([0-9]+) - ([0-9]+) - ([0-9]+) - \\1 - \\2 - \\3(_[0-9]+)?$",
      "Role of AES schemes on the grasslands biodiversity - \\1-\\2-\\3 - \\1-\\2-\\3\\4"
      )
    )
```

## Join extent and biodiversity data

```{r}
biodiv_extent <- RF_data %>% 
  # inner_join only joins records present in both
  # some are present in RF_data but not in biodiv_perc 
  # (e.g. DK presence-absence data)
  # some are present in biodiv_perc but not in RF_data 
  # (e.g. those not belonging to Q, R, S, T)
  inner_join(biodiv_perc, by = c("plotID" = "time_seriesID"),
            relationship = "many-to-many")
```

## Check plots with extent from Mahalanobis and not from RF

```{r}
plotIDs_extent_Maha <- readRDS(here("data", "clean", "biodiv_extent.rds")) %>% 
  distinct(plotID)
plotIDs_extent_RF <- bind_rows(biodiv_extent %>% distinct(plotID))
# Plots with extent from Mahalanobis and not from RF
plotIDs_extent_Maha %>%
  anti_join(
    plotIDs_extent_RF,
    by = "plotID"
  )
```

Only one, fine!

## Check if years match

```{r}
# Number of plots
biodiv_extent %>% distinct(plotID) %>% nrow()
# Number of plots where first and last year match --> USE THIS SO FAR!
biodiv_extent %>% filter(first_year.x == year_start & last_year == year_end) %>% 
  distinct(plotID) %>% nrow()
# Number of plots where first year does not match, but last year matches
biodiv_extent %>% filter(first_year.x != year_start & last_year == year_end) %>% 
  distinct(plotID) %>% nrow()
# Number of plots where first year matches, but last year does not
biodiv_extent %>% filter(first_year.x == year_start & last_year != year_end) %>% 
  distinct(plotID) %>% nrow()
# Number of plots where neither first nor last year match
biodiv_extent %>% filter(first_year.x != year_start & last_year != year_end) %>% 
  distinct(plotID) %>% nrow()
```

In some cases, they do not match because the first or last observation in some time series did not have data on functional and/or phylogenetic diversity metrics (cf. Stephan's email). So far, we use only time series where first and last year match.

```{r}
biodiv_extent <- biodiv_extent %>%
  filter(first_year.x == year_start & last_year == year_end)
```

# Numbers of plots

```{r}
# Total
biodiv_extent %>% distinct(plotID) %>% nrow()
# With constant habitat
biodiv_extent %>% filter(eunis_cambio == F) %>% distinct(plotID) %>% nrow()
# with habitat change
biodiv_extent %>% filter(eunis_cambio == T) %>% distinct(plotID) %>% nrow()
```

# TBD: Map of plots in bioregions

# Count of plots in bioregions

```{r}
biodiv_extent %>% dplyr::select(plotID, biogeo) %>% 
  distinct() %>% count(biogeo) %>% arrange(n)
```

# Distributions

## Initial biodiversity values

### Plots with constant habitat

```{r}
# Histograms
biodiv_extent %>%
  filter(eunis_cambio == F) %>%
  group_by(metric) %>%
  mutate(
    q01 = quantile(initial_div_value, 0.01, na.rm = TRUE),
    q99 = quantile(initial_div_value, 0.99, na.rm = TRUE)
  ) %>%
  filter(initial_div_value >= q01, initial_div_value <= q99) %>%
  ggplot(aes(initial_div_value)) +
  geom_histogram(bins = 30, color = "black", fill = "grey") +
  facet_wrap(~metric, scales = "free", ncol = 6) +
  theme_bw() + xlab("Initial diversity value")
```

### Plots with habitat change

```{r}
# Histograms
biodiv_extent %>%
  filter(eunis_cambio == T) %>%
  group_by(metric) %>%
  mutate(
    q01 = quantile(initial_div_value, 0.01, na.rm = TRUE),
    q99 = quantile(initial_div_value, 0.99, na.rm = TRUE)
  ) %>%
  filter(initial_div_value >= q01, initial_div_value <= q99) %>%
  ggplot(aes(initial_div_value)) +
  geom_histogram(bins = 30, color = "black", fill = "grey") +
  facet_wrap(~metric, scales = "free", ncol = 6) +
  theme_bw() + xlab("Initial diversity value")
```

## Biodiversity change

### Plots with constant habitat

```{r}
# Histograms
biodiv_extent %>%
  filter(eunis_cambio == F) %>%
  group_by(metric) %>%
  mutate(
    q01 = quantile(slope, 0.01, na.rm = TRUE),
    q99 = quantile(slope, 0.99, na.rm = TRUE)
  ) %>%
  filter(slope >= q01, slope <= q99) %>%
  ggplot(aes(slope)) +
  geom_histogram(bins = 30, color = "black", fill = "grey") +
  facet_wrap(~metric, scales = "free", ncol = 6) +
  theme_bw() + xlab("Regression slope (annual percentage change)")
```

### Plots with habitat change

```{r}
# Histograms
biodiv_extent %>%
  filter(eunis_cambio == F) %>%
  group_by(metric) %>%
  mutate(
    q01 = quantile(slope, 0.01, na.rm = TRUE),
    q99 = quantile(slope, 0.99, na.rm = TRUE)
  ) %>%
  filter(slope >= q01, slope <= q99) %>%
  ggplot(aes(slope)) +
  geom_histogram(bins = 30, color = "black", fill = "grey") +
  facet_wrap(~metric, scales = "free", ncol = 6) +
  theme_bw() + xlab("Regression slope (annual percentage change)")
```

## Habitat extent change

```{r}
# Histograms
biodiv_extent %>%
  distinct(plotID, eunis_cambio, delta_focal_ha, delta_focal_pct) %>%
  pivot_longer(cols = c(delta_focal_ha:delta_focal_pct), names_to = "metric",
               values_to = "value") %>%
  ggplot(aes(value)) +
  geom_histogram(bins = 30, color = "black", fill = "grey") +
  facet_wrap(~ metric + eunis_cambio, scales = "free") +
  theme_bw() + xlab("Change in extent")
```

# Correlations

## Correlations among initial biodiversity values

```{r}
biodiv_wide <- biodiv_extent %>%
  distinct(plotID, metric, initial_div_value) %>%
  pivot_wider(names_from = metric, values_from = initial_div_value)
```

```{r}
cor_mat_init <- biodiv_wide %>%
  dplyr::select(-plotID) %>%
  cor(use = "pairwise.complete.obs")
corrplot(cor_mat_init, method = "color", type = "upper", addCoef.col = "black",
         number.cex = 0.6, tl.col = "black", order = "hclust")
```

## Correlations among biodiversity change metrics

```{r}
biodiv_change_wide <- biodiv_extent %>%
  distinct(plotID, metric, slope) %>%
  pivot_wider(names_from = metric, values_from = slope)
```

```{r}
cor_mat_change <- biodiv_change_wide %>%
  dplyr::select(-plotID) %>%
  cor(use = "pairwise.complete.obs")
corrplot(cor_mat_change, method = "color", type = "upper", addCoef.col = "black",
         number.cex = 0.6, tl.col = "black", order = "hclust")
```

# Models change in EXTENT

## Effect of change in extent on change in diversity

```{r}
models_change <- biodiv_extent %>%
  group_by(metric) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(slope ~ delta_focal_pct, data = .x)),
    coef = map(model, ~ tidy(.x) %>%
                 filter(term == "delta_focal_pct")),
    fit = map(model, glance),
    estimate = map_dbl(coef, "estimate"),
    p.value  = map_dbl(coef, "p.value"),
    r2       = map_dbl(fit, "r.squared"),
    n        = map_int(data, nrow)
  ) %>%
  dplyr::select(metric, model, estimate, p.value, r2, n)
```

Check performance of models:

```{r}
models_change %>%
  pull(model) %>%
  walk(~ print(check_model(.x)))
```

```{r}
models_change %>% filter(p.value < 0.05) %>% arrange(metric)
```

```{r}
models_change %>% filter(p.value < 0.05) %>% arrange(metric) %>%
  pull(model) %>%
  walk(~ print(check_model(.x)))
```


```{r}
plot_data_change <- biodiv_extent %>%
  semi_join(
    models_change %>%
      filter(p.value < 0.05) %>%
      dplyr::select(metric),
    by = c("metric")
  ) %>%
  left_join(
    models_change %>%
      dplyr::select(metric, estimate, p.value, r2, n) %>%
      rename(n_obs = n),
    by = c("metric")
  )
```

```{r}
make_plot_change <- function(df){

  biodiv_name <- unique(df$metric)
  estimate <- unique(df$estimate)
  pval <- unique(df$p.value)
  r2 <- unique(df$r2)
  n_obs <- unique(df$n_obs)

  ggplot(df, aes(x = delta_focal_pct, y = slope)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "lm", colour = "blue", fill = "blue", alpha = 0.15) +
    geom_smooth(method = MASS::rlm, colour = "red", se = FALSE) +
    annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.1, size = 4,
             label = sprintf("β = %.3f\nR² = %.3f\np = %.3g\nn obs = %d",
                             estimate, r2, pval, n_obs)) +
    labs(title = paste0(biodiv_name), 
         x = "Change in habitat extent (%)",
         y = paste0("Change in ", biodiv_name), colour = NULL) +
    theme_bw()
}
```

```{r message=FALSE}
plot_data_change %>%
  group_by(metric) %>%
  group_split() %>%
  map(make_plot_change)
```

## Effect of change in extent for each habitat type

```{r}
models_change_hab <- biodiv_extent %>%
  group_by(metric, eunis1) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(slope ~ delta_focal_pct, data = .x)),
    coef = map(model, ~ tidy(.x) %>%
                 filter(term == "delta_focal_pct")),
    fit = map(model, glance),
    estimate = map_dbl(coef, "estimate"),
    p.value  = map_dbl(coef, "p.value"),
    r2       = map_dbl(fit, "r.squared"),
    n        = map_int(data, nrow)
  ) %>%
  dplyr::select(metric, eunis1, model, estimate, p.value, r2, n)
```

### TBD? Check performance of models

```{r}
models_change_hab %>% filter(p.value < 0.05) %>% arrange(metric, eunis1)
```

```{r}
plot_data_change_hab <- biodiv_extent %>%
  semi_join(
    models_change_hab %>%
      dplyr::filter(p.value < 0.05),
    by = c("metric", "eunis1")
  ) %>%
  left_join(
    models_change_hab %>%
      dplyr::select(metric, eunis1, estimate, p.value, r2, n) %>%
      rename(n_obs = n),
    by = c("metric", "eunis1")
  )
```

```{r}
make_plot_change_hab <- function(df){

  biodiv_name <- unique(df$metric)
  EUNIS <- unique(df$eunis1)
  estimate <- unique(df$estimate)
  pval <- unique(df$p.value)
  r2 <- unique(df$r2)
  n_obs <- unique(df$n_obs)

  ggplot(df, aes(x = delta_focal_pct, y = slope)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "lm", colour = "blue", fill = "blue", alpha = 0.15) +
    geom_smooth(method = MASS::rlm, colour = "red", se = FALSE) +
    annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.1, size = 4,
             label = sprintf("β = %.3f\nR² = %.3f\np = %.3g\nn obs = %d",
                             estimate, r2, pval, n_obs)) +
    labs(title = paste0(biodiv_name, " | EUNIS = ", EUNIS),
         x = "Change in habitat extent (%)",
         y = paste0("Change in ", biodiv_name)) +
    theme_bw()
}
```

```{r message=FALSE}
plot_data_change_hab %>%
  group_by(metric, eunis1) %>%
  group_split() %>%
  map(make_plot_change_hab)
```

## Effect of change in extent for each trajectory type

```{r}
models_change_traj <- biodiv_extent %>%
  group_by(metric, Trajectory) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(slope ~ delta_focal_pct, data = .x)),
    coef = map(model, ~ tidy(.x) %>%
                 filter(term == "delta_focal_pct")),
    fit = map(model, glance),
    estimate = map_dbl(coef, "estimate"),
    p.value  = map_dbl(coef, "p.value"),
    r2       = map_dbl(fit, "r.squared"),
    n        = map_int(data, nrow)
  ) %>%
  dplyr::select(metric, Trajectory, model, estimate, p.value, r2, n)
```

```{r}
models_change_traj %>% filter(p.value < 0.05) %>% arrange(metric, Trajectory)
```

```{r}
plot_data_change_traj <- biodiv_extent %>%
  semi_join(
    models_change_traj %>%
      dplyr::filter(p.value < 0.05),
    by = c("metric", "Trajectory")
  ) %>%
  left_join(
    models_change_traj %>%
      dplyr::select(metric, Trajectory, estimate, p.value, r2, n) %>%
      rename(n_obs = n),
    by = c("metric", "Trajectory")
  )
```

```{r}
make_plot_change_traj <- function(df){

  biodiv_name <- unique(df$metric)
  traj <- unique(df$Trajectory)
  estimate <- unique(df$estimate)
  pval <- unique(df$p.value)
  r2 <- unique(df$r2)
  n_obs <- unique(df$n_obs)

  ggplot(df, aes(x = delta_focal_pct, y = slope)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "lm", colour = "blue", fill = "blue", alpha = 0.15) +
    geom_smooth(method = MASS::rlm, colour = "red", se = FALSE) +
    annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.1, size = 4,
             label = sprintf("β = %.3f\nR² = %.3f\np = %.3g\nn obs = %d",
                             estimate, r2, pval, n_obs)) +
    labs(title = paste0(biodiv_name, " | Trajectory = ", traj),
         x = "Change in habitat extent (%)", 
         y = paste0("Change in ", biodiv_name)) +
    theme_bw()
}
```

```{r message=FALSE}
plot_data_change_traj %>%
  group_by(metric, Trajectory) %>%
  group_split() %>%
  map(make_plot_change_traj)
```

# Models change in FUNCTIONAL METRICS

## Effect of change in functional metrics on change in diversity

```{r}
models_change_funct <- biodiv_extent %>%
  pivot_longer(cols = c(ndvi_resurvey, ndmi_resurvey, ndwi_resurvey, 
                        savi_resurvey),
               names_to = "rs_index", values_to = "rs_slope") %>%
  group_by(metric, rs_index) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(slope ~ rs_slope, data = .x)),
    coef = map(model, ~ tidy(.x) %>%
                 filter(term == "rs_slope")),
    fit = map(model, glance),
    estimate = map_dbl(coef, "estimate"),
    p.value  = map_dbl(coef, "p.value"),
    r2       = map_dbl(fit, "r.squared"),
    n        = map_int(data, nrow)
  ) %>%
  dplyr::select(metric, rs_index, model, estimate, p.value, r2, n)
```

```{r}
models_change_funct %>% filter(p.value < 0.05) %>% arrange(metric, rs_index)
```

```{r}
plot_data_change_funct <- biodiv_extent %>%
  pivot_longer(cols = c(ndvi_resurvey, ndmi_resurvey, ndwi_resurvey, 
                        savi_resurvey),
               names_to = "rs_index", values_to = "rs_slope") %>%
  semi_join(models_change_funct %>% filter(p.value < 0.05),
            by = c("metric", "rs_index")) %>%
  left_join(models_change_funct %>%
              dplyr::select(metric, rs_index, estimate, p.value, r2, n) %>%
              rename(n_obs = n),
            by = c("metric", "rs_index")
            )
```

```{r}
make_plot_change_funct <- function(df){
  
  biodiv_name <- unique(df$metric)
  rs_index <- unique(df$rs_index)
  estimate <- unique(df$estimate)
  pval <- unique(df$p.value)
  r2 <- unique(df$r2)
  n_obs <- unique(df$n_obs)
  
  ggplot(df, aes(x = rs_slope, y = slope)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "lm", colour = "blue", fill = "blue", alpha = 0.15) +
    geom_smooth(method = MASS::rlm, colour = "red", se = FALSE) +
    annotate("text", x = Inf, y = Inf, hjust = 1.3, vjust = 1.1,size = 4,
             label = sprintf("β = %.3f\nR² = %.3f\np = %.3g\nn = %d",
                             estimate, r2, pval, n_obs)) +
    labs(title = paste0(biodiv_name, " | ", rs_index),
         x = paste0(rs_index, " Sen's slope"),
         y = paste0("Change in ", biodiv_name)) +
    theme_bw()
  }
```

```{r message=FALSE, warning=FALSE}
plot_data_change_funct %>%
  group_by(metric, rs_index) %>%
  group_split() %>%
  map(make_plot_change_funct)
```

## Effect of change in functional metrics for each habitat type

```{r}
models_change_funct_hab <- biodiv_extent %>%
  pivot_longer(cols = c(ndvi_resurvey, ndmi_resurvey, ndwi_resurvey, 
                        savi_resurvey),
               names_to = "rs_index", values_to = "rs_slope") %>%
  group_by(metric, rs_index, eunis1) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(slope ~ rs_slope, data = .x)),
    coef = map(model, ~ tidy(.x) %>%
                 filter(term == "rs_slope")),
    fit = map(model, glance),
    estimate = map_dbl(coef, "estimate"),
    p.value  = map_dbl(coef, "p.value"),
    r2       = map_dbl(fit, "r.squared"),
    n        = map_int(data, nrow)
  ) %>%
  dplyr::select(metric, rs_index, eunis1, model, estimate, p.value, r2, n)
```

```{r}
models_change_funct_hab %>% filter(p.value < 0.05 & r2 > 0.1) %>% 
  arrange(metric, rs_index, eunis1)
```

```{r}
plot_data_change_funct_hab <- biodiv_extent %>%
  pivot_longer(cols = c(ndvi_resurvey, ndmi_resurvey, ndwi_resurvey, 
                        savi_resurvey),
               names_to = "rs_index", values_to = "rs_slope") %>%
  semi_join(
    models_change_funct_hab %>% dplyr::filter(p.value < 0.05 & r2 > 0.1),
    by = c("metric", "rs_index", "eunis1")
  ) %>%
  left_join(
    models_change_funct_hab %>%
      dplyr::select(metric, rs_index, eunis1, estimate, p.value, r2, n) %>%
      rename(n_obs = n),
    by = c("metric", "rs_index", "eunis1")
  )
```

```{r}
make_plot_change_funct_hab <- function(df){

  biodiv_name <- unique(df$metric)
  rs_index <- unique(df$rs_index)
  EUNIS <- unique(df$eunis1)
  estimate <- unique(df$estimate)
  pval <- unique(df$p.value)
  r2 <- unique(df$r2)
  n_obs <- unique(df$n_obs)

  ggplot(df, aes(x = rs_slope, y = slope)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "lm", colour = "blue", fill = "blue", alpha = 0.15) +
    geom_smooth(method = MASS::rlm, colour = "red", se = FALSE) +
    annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.1, size = 4,
             label = sprintf("β = %.3f\nR² = %.3f\np = %.3g\nn obs = %d",
                             estimate, r2, pval, n_obs)) +
    labs(title = paste0(biodiv_name, " | EUNIS = ", EUNIS),
         x = paste0(rs_index, " Sen's slope"),
         y = paste0("Change in ", biodiv_name)) +
    theme_bw()
}
```

```{r message=FALSE}
plot_data_change_funct_hab %>%
  group_by(metric, rs_index, eunis1) %>%
  group_split() %>%
  map(make_plot_change_funct_hab)
```

## Effect of change in functional metrics for each trajectory type

```{r}
models_change_funct_traj <- biodiv_extent %>%
  pivot_longer(cols = c(ndvi_resurvey, ndmi_resurvey, ndwi_resurvey, 
                        savi_resurvey),
               names_to = "rs_index", values_to = "rs_slope") %>%
  group_by(metric, rs_index, Trajectory) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(slope ~ rs_slope, data = .x)),
    coef = map(model, ~ tidy(.x) %>%
                 filter(term == "rs_slope")),
    fit = map(model, glance),
    estimate = map_dbl(coef, "estimate"),
    p.value  = map_dbl(coef, "p.value"),
    r2       = map_dbl(fit, "r.squared"),
    n        = map_int(data, nrow)
  ) %>%
  dplyr::select(metric, rs_index, Trajectory, model, estimate, p.value, r2, n)
```

```{r}
models_change_funct_traj %>% filter(p.value < 0.05 & r2 > 0.2) %>% 
  arrange(metric, rs_index, Trajectory)
```

```{r}
plot_data_change_funct_traj <- biodiv_extent %>%
  pivot_longer(cols = c(ndvi_resurvey, ndmi_resurvey, ndwi_resurvey, 
                        savi_resurvey),
               names_to = "rs_index", values_to = "rs_slope") %>%
  semi_join(
    models_change_funct_traj %>% dplyr::filter(p.value < 0.05 & r2 > 0.2),
    by = c("metric", "rs_index", "Trajectory")
  ) %>%
  left_join(
    models_change_funct_traj %>%
      dplyr::select(metric, rs_index, Trajectory, estimate, p.value, r2, n) %>%
      rename(n_obs = n),
    by = c("metric", "rs_index", "Trajectory")
  )
```

```{r}
make_plot_change_funct_traj <- function(df){

  biodiv_name <- unique(df$metric)
  rs_index <- unique(df$rs_index)
  Trajectory <- unique(df$Trajectory)
  estimate <- unique(df$estimate)
  pval <- unique(df$p.value)
  r2 <- unique(df$r2)
  n_obs <- unique(df$n_obs)

  ggplot(df, aes(x = rs_slope, y = slope)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "lm", colour = "blue", fill = "blue", alpha = 0.15) +
    geom_smooth(method = MASS::rlm, colour = "red", se = FALSE) +
    annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.1, size = 4,
             label = sprintf("β = %.3f\nR² = %.3f\np = %.3g\nn obs = %d",
                             estimate, r2, pval, n_obs)) +
    labs(title = paste0(biodiv_name, " | Trajectory = ", Trajectory),
         x = paste0(rs_index, " Sen's slope"),
         y = paste0("Change in ", biodiv_name)) +
    theme_bw()
}
```

```{r message=FALSE}
plot_data_change_funct_traj %>%
  group_by(metric, rs_index, Trajectory) %>%
  group_split() %>%
  map(make_plot_change_funct_traj)
```

# TBD: Try other model types instead of lm() for some variables

# Models change in EXTENT without OUTLIERS

Removing upper and lower 1% quantiles of biodiversity change values (slopes) and re-running the models.

## Effect of change in extent on change in diversity

```{r}
models_change_noout <- biodiv_extent %>%
  group_by(metric) %>%
  # Define quantiles
  mutate(
    q01 = quantile(slope, 0.01, na.rm = TRUE),
    q99 = quantile(slope, 0.99, na.rm = TRUE)
  ) %>%
  # Reemove outliers
  filter(slope >= q01, slope <= q99) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(slope ~ delta_focal_pct, data = .x)),
    coef = map(model, ~ tidy(.x) %>%
                 filter(term == "delta_focal_pct")),
    fit = map(model, glance),
    estimate = map_dbl(coef, "estimate"),
    p.value  = map_dbl(coef, "p.value"),
    r2       = map_dbl(fit, "r.squared"),
    n        = map_int(data, nrow)
  ) %>%
  dplyr::select(metric, model, estimate, p.value, r2, n)
```

```{r}
models_change_noout %>% filter(p.value < 0.05) %>% arrange(metric)
```

```{r}
plot_data_change_noout <- biodiv_extent %>%
  group_by(metric) %>%
  # Define quantiles
  mutate(
    q01 = quantile(slope, 0.01, na.rm = TRUE),
    q99 = quantile(slope, 0.99, na.rm = TRUE)
  ) %>%
  # Reemove outliers
  filter(slope >= q01, slope <= q99) %>%
  ungroup() %>%
  semi_join(
    models_change_noout %>%
      filter(p.value < 0.05) %>%
      dplyr::select(metric),
    by = c("metric")
  ) %>%
  left_join(
    models_change_noout %>%
      dplyr::select(metric, estimate, p.value, r2, n) %>%
      rename(n_obs = n),
    by = c("metric")
  )
```

```{r message=FALSE}
plot_data_change_noout %>%
  group_by(metric) %>%
  group_split() %>%
  map(make_plot_change)
```

## Effect of change in extent for each habitat type

```{r}
models_change_hab_noout <- biodiv_extent %>%
  group_by(metric) %>%
  # Define quantiles
  mutate(
    q01 = quantile(slope, 0.01, na.rm = TRUE),
    q99 = quantile(slope, 0.99, na.rm = TRUE)
  ) %>%
  # Reemove outliers
  filter(slope >= q01, slope <= q99) %>%
  group_by(metric, eunis1) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(slope ~ delta_focal_pct, data = .x)),
    coef = map(model, ~ tidy(.x) %>%
                 filter(term == "delta_focal_pct")),
    fit = map(model, glance),
    estimate = map_dbl(coef, "estimate"),
    p.value  = map_dbl(coef, "p.value"),
    r2       = map_dbl(fit, "r.squared"),
    n        = map_int(data, nrow)
  ) %>%
  dplyr::select(metric, eunis1, model, estimate, p.value, r2, n)
```

```{r}
models_change_hab_noout %>% filter(p.value < 0.05) %>% arrange(metric, eunis1)
```

```{r}
plot_data_change_hab_noout <- biodiv_extent %>%
  group_by(metric) %>%
  # Define quantiles
  mutate(
    q01 = quantile(slope, 0.01, na.rm = TRUE),
    q99 = quantile(slope, 0.99, na.rm = TRUE)
  ) %>%
  # Reemove outliers
  filter(slope >= q01, slope <= q99) %>%
  ungroup() %>%
  semi_join(
    models_change_hab_noout %>%
      dplyr::filter(p.value < 0.05),
    by = c("metric", "eunis1")
  ) %>%
  left_join(
    models_change_hab_noout %>%
      dplyr::select(metric, eunis1, estimate, p.value, r2, n) %>%
      rename(n_obs = n),
    by = c("metric", "eunis1")
  )
```

```{r message=FALSE}
plot_data_change_hab_noout %>%
  group_by(metric, eunis1) %>%
  group_split() %>%
  map(make_plot_change_hab)
```

## Effect of change in extent for each trajectory type

```{r}
models_change_traj_noout <- biodiv_extent %>%
  group_by(metric) %>%
  # Define quantiles
  mutate(
    q01 = quantile(slope, 0.01, na.rm = TRUE),
    q99 = quantile(slope, 0.99, na.rm = TRUE)
  ) %>%
  # Reemove outliers
  filter(slope >= q01, slope <= q99) %>%
  group_by(metric, Trajectory) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(slope ~ delta_focal_pct, data = .x)),
    coef = map(model, ~ tidy(.x) %>%
                 filter(term == "delta_focal_pct")),
    fit = map(model, glance),
    estimate = map_dbl(coef, "estimate"),
    p.value  = map_dbl(coef, "p.value"),
    r2       = map_dbl(fit, "r.squared"),
    n        = map_int(data, nrow)
  ) %>%
  dplyr::select(metric, Trajectory, model, estimate, p.value, r2, n)
```

```{r}
models_change_traj_noout %>% filter(p.value < 0.05) %>% arrange(metric, Trajectory)
```

```{r}
plot_data_change_traj_noout <- biodiv_extent %>%
  group_by(metric) %>%
  # Define quantiles
  mutate(
    q01 = quantile(slope, 0.01, na.rm = TRUE),
    q99 = quantile(slope, 0.99, na.rm = TRUE)
  ) %>%
  # Reemove outliers
  filter(slope >= q01, slope <= q99) %>%
  ungroup() %>%
  semi_join(
    models_change_traj_noout %>%
      dplyr::filter(p.value < 0.05),
    by = c("metric", "Trajectory")
  ) %>%
  left_join(
    models_change_traj_noout %>%
      dplyr::select(metric, Trajectory, estimate, p.value, r2, n) %>%
      rename(n_obs = n),
    by = c("metric", "Trajectory")
  )
```

```{r message=FALSE}
plot_data_change_traj_noout %>%
  group_by(metric, Trajectory) %>%
  group_split() %>%
  map(make_plot_change_traj)
```

# Models change in FUNCTIONAL METRICS WITHOUT OUTLIERS

## Effect of change in functional metrics on change in diversity

```{r}
models_change_funct_noout <- biodiv_extent %>%
  group_by(metric) %>%
  # Define quantiles
  mutate(
    q01 = quantile(slope, 0.01, na.rm = TRUE),
    q99 = quantile(slope, 0.99, na.rm = TRUE)
  ) %>%
  # Reemove outliers
  filter(slope >= q01, slope <= q99) %>%
  pivot_longer(cols = c(ndvi_resurvey, ndmi_resurvey, ndwi_resurvey, 
                        savi_resurvey),
               names_to = "rs_index", values_to = "rs_slope") %>%
  group_by(metric, rs_index) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(slope ~ rs_slope, data = .x)),
    coef = map(model, ~ tidy(.x) %>%
                 filter(term == "rs_slope")),
    fit = map(model, glance),
    estimate = map_dbl(coef, "estimate"),
    p.value  = map_dbl(coef, "p.value"),
    r2       = map_dbl(fit, "r.squared"),
    n        = map_int(data, nrow)
  ) %>%
  dplyr::select(metric, rs_index, model, estimate, p.value, r2, n)
```

```{r}
models_change_funct_noout %>% filter(p.value < 0.05) %>% arrange(metric, rs_index)
```

```{r}
plot_data_change_funct_noout <- biodiv_extent %>%
  group_by(metric) %>%
  # Define quantiles
  mutate(
    q01 = quantile(slope, 0.01, na.rm = TRUE),
    q99 = quantile(slope, 0.99, na.rm = TRUE)
  ) %>%
  # Reemove outliers
  filter(slope >= q01, slope <= q99) %>%
  pivot_longer(cols = c(ndvi_resurvey, ndmi_resurvey, ndwi_resurvey, 
                        savi_resurvey),
               names_to = "rs_index", values_to = "rs_slope") %>%
  semi_join(models_change_funct_noout %>% filter(p.value < 0.05),
            by = c("metric", "rs_index")) %>%
  left_join(models_change_funct_noout %>%
              dplyr::select(metric, rs_index, estimate, p.value, r2, n) %>%
              rename(n_obs = n),
            by = c("metric", "rs_index")
            )
```

```{r message=FALSE, warning=FALSE}
plot_data_change_funct_noout %>%
  group_by(metric, rs_index) %>%
  group_split() %>%
  map(make_plot_change_funct)
```

## Effect of change in functional metrics for each habitat type

```{r}
models_change_funct_hab_noout <- biodiv_extent %>%
  group_by(metric) %>%
  # Define quantiles
  mutate(
    q01 = quantile(slope, 0.01, na.rm = TRUE),
    q99 = quantile(slope, 0.99, na.rm = TRUE)
  ) %>%
  # Reemove outliers
  filter(slope >= q01, slope <= q99) %>%
  pivot_longer(cols = c(ndvi_resurvey, ndmi_resurvey, ndwi_resurvey, 
                        savi_resurvey),
               names_to = "rs_index", values_to = "rs_slope") %>%
  group_by(metric, rs_index, eunis1) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(slope ~ rs_slope, data = .x)),
    coef = map(model, ~ tidy(.x) %>%
                 filter(term == "rs_slope")),
    fit = map(model, glance),
    estimate = map_dbl(coef, "estimate"),
    p.value  = map_dbl(coef, "p.value"),
    r2       = map_dbl(fit, "r.squared"),
    n        = map_int(data, nrow)
  ) %>%
  dplyr::select(metric, rs_index, eunis1, model, estimate, p.value, r2, n)
```

```{r}
models_change_funct_hab_noout %>% filter(p.value < 0.05 & r2 > 0.05) %>% 
  arrange(metric, rs_index, eunis1)
```

```{r}
plot_data_change_funct_hab_noout <- biodiv_extent %>%
  group_by(metric) %>%
  # Define quantiles
  mutate(
    q01 = quantile(slope, 0.01, na.rm = TRUE),
    q99 = quantile(slope, 0.99, na.rm = TRUE)
  ) %>%
  # Reemove outliers
  filter(slope >= q01, slope <= q99) %>%
  pivot_longer(cols = c(ndvi_resurvey, ndmi_resurvey, ndwi_resurvey, 
                        savi_resurvey),
               names_to = "rs_index", values_to = "rs_slope") %>%
  semi_join(
    models_change_funct_hab_noout %>% dplyr::filter(p.value < 0.05 & r2 > 0.05),
    by = c("metric", "rs_index", "eunis1")
  ) %>%
  left_join(
    models_change_funct_hab_noout %>%
      dplyr::select(metric, rs_index, eunis1, estimate, p.value, r2, n) %>%
      rename(n_obs = n),
    by = c("metric", "rs_index", "eunis1")
  )
```

```{r message=FALSE}
plot_data_change_funct_hab_noout %>%
  group_by(metric, rs_index, eunis1) %>%
  group_split() %>%
  map(make_plot_change_funct_hab)
```

## Effect of change in functional metrics for each trajectory type

```{r}
models_change_funct_traj_noout <- biodiv_extent %>%
  group_by(metric) %>%
  # Define quantiles
  mutate(
    q01 = quantile(slope, 0.01, na.rm = TRUE),
    q99 = quantile(slope, 0.99, na.rm = TRUE)
  ) %>%
  # Reemove outliers
  filter(slope >= q01, slope <= q99) %>%
  pivot_longer(cols = c(ndvi_resurvey, ndmi_resurvey, ndwi_resurvey, 
                        savi_resurvey),
               names_to = "rs_index", values_to = "rs_slope") %>%
  group_by(metric, rs_index, Trajectory) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(slope ~ rs_slope, data = .x)),
    coef = map(model, ~ tidy(.x) %>%
                 filter(term == "rs_slope")),
    fit = map(model, glance),
    estimate = map_dbl(coef, "estimate"),
    p.value  = map_dbl(coef, "p.value"),
    r2       = map_dbl(fit, "r.squared"),
    n        = map_int(data, nrow)
  ) %>%
  dplyr::select(metric, rs_index, Trajectory, model, estimate, p.value, r2, n)
```

```{r}
models_change_funct_traj_noout %>% filter(p.value < 0.05 & r2 > 0.05) %>% 
  arrange(metric, rs_index, Trajectory)
```

```{r}
plot_data_change_funct_traj_noout <- biodiv_extent %>%
  group_by(metric) %>%
  # Define quantiles
  mutate(
    q01 = quantile(slope, 0.01, na.rm = TRUE),
    q99 = quantile(slope, 0.99, na.rm = TRUE)
  ) %>%
  # Reemove outliers
  filter(slope >= q01, slope <= q99) %>%
  pivot_longer(cols = c(ndvi_resurvey, ndmi_resurvey, ndwi_resurvey, 
                        savi_resurvey),
               names_to = "rs_index", values_to = "rs_slope") %>%
  semi_join(
    models_change_funct_traj_noout %>% dplyr::filter(p.value < 0.05 & r2 > 0.05),
    by = c("metric", "rs_index", "Trajectory")
  ) %>%
  left_join(
    models_change_funct_traj_noout %>%
      dplyr::select(metric, rs_index, Trajectory, estimate, p.value, r2, n) %>%
      rename(n_obs = n),
    by = c("metric", "rs_index", "Trajectory")
  )
```

```{r message=FALSE}
plot_data_change_funct_traj_noout %>%
  group_by(metric, rs_index, Trajectory) %>%
  group_split() %>%
  map(make_plot_change_funct_traj)
```



# RDA slopes CWMs ~ slopes functional metrics

## All

```{r}
cwm_wide <- biodiv_extent %>%
  filter(str_starts(metric, "CWM_")) %>%
  dplyr::select(plotID, ndvi_resurvey, ndmi_resurvey, ndwi_resurvey, 
                savi_resurvey, metric, slope) %>% 
  distinct() %>%
  pivot_wider(names_from = metric, values_from = slope) %>%
  # Remove NAs
  drop_na()
```

Response matrix:

```{r}
Y <- cwm_wide %>%
  dplyr::select(CWM_PlantHeight, CWM_RootingDepth, CWM_LeafCNratio, 
                CWM_LeafArea, CWM_SLA) %>%
  scale()
```

Explanatory variables:

```{r}
X <- cwm_wide %>%
  dplyr::select(ndvi_resurvey, ndmi_resurvey, ndwi_resurvey, savi_resurvey)
```

RDA:

```{r}
library(vegan)

mod_rda <- rda(Y ~ ndvi_resurvey + ndmi_resurvey + ndwi_resurvey + 
                 savi_resurvey, data = X)
```

```{r}
anova(mod_rda)
anova(mod_rda, by = "axis")
anova(mod_rda, by = "term")
```

```{r}
RsquareAdj(mod_rda)
```

```{r}
library(ggvegan)

autoplot(mod_rda, layers = c("species", "biplot"))
```

## By habitat

```{r}
library(tidyverse)
library(vegan)

rda_habitat <- biodiv_extent %>%
  filter(str_starts(metric, "CWM_")) %>%
  dplyr::select(
    plotID, eunis1,
    ndvi_resurvey, ndmi_resurvey,
    ndwi_resurvey, savi_resurvey,
    metric, slope
  ) %>%
  pivot_wider(
    names_from = metric,
    values_from = slope
  ) %>%
  drop_na() %>%
  group_by(eunis1) %>%
  nest() %>%
  mutate(
    rda = map(
      data,
      ~{
        Y <- .x %>%
          dplyr::select(
            CWM_PlantHeight,
            CWM_RootingDepth,
            CWM_LeafCNratio,
            CWM_LeafArea,
            CWM_SLA
          ) %>%
          scale()

        X <- .x %>%
          dplyr::select(
            ndvi_resurvey,
            ndmi_resurvey,
            ndwi_resurvey,
            savi_resurvey
          )

        rda(Y ~ ., data = X)
      }
    )
  )
```

```{r}
anova(rda_habitat$rda[[1]])
anova(rda_habitat$rda[[2]])
anova(rda_habitat$rda[[3]])
anova(rda_habitat$rda[[4]])
anova(rda_habitat$rda[[5]])

anova(rda_habitat$rda[[1]], by = "axis")
anova(rda_habitat$rda[[2]], by = "axis")
anova(rda_habitat$rda[[3]], by = "axis")
anova(rda_habitat$rda[[4]], by = "axis")
anova(rda_habitat$rda[[5]], by = "axis")

anova(rda_habitat$rda[[1]], by = "term")
anova(rda_habitat$rda[[2]], by = "term")
anova(rda_habitat$rda[[3]], by = "term")
anova(rda_habitat$rda[[4]], by = "term")
anova(rda_habitat$rda[[5]], by = "term")
```

```{r}
RsquareAdj(rda_habitat$rda[[1]])
RsquareAdj(rda_habitat$rda[[2]])
RsquareAdj(rda_habitat$rda[[3]])
RsquareAdj(rda_habitat$rda[[4]])
RsquareAdj(rda_habitat$rda[[5]])
```

```{r}
autoplot(rda_habitat$rda[[1]], layers = c("species", "biplot"))
autoplot(rda_habitat$rda[[2]], layers = c("species", "biplot"))
autoplot(rda_habitat$rda[[3]], layers = c("species", "biplot"))
autoplot(rda_habitat$rda[[4]], layers = c("species", "biplot"))
autoplot(rda_habitat$rda[[5]], layers = c("species", "biplot"))
```


# Session info

```{r}
sessionInfo()
```

