---
title: "MOTIVATE mapping - Calculating habitat extent from Mahalanobis rasters"
author: "Alicia Valdés"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: 4
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  word_document:
    toc: yes
    toc_depth: '4'
editor_options:
  chunk_output_type: inline
---

This notebook calculates habitat extent from Mahalanobis rasters.

# Load libraries

```{r}
library(terra)
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(sf)
library(readr)
library(here)
library(exactextractr)
library(furrr)
library(tmap)
```

# Load data

Load points from ReSurvey Europe observations from validated plots. Some (most) of these observations are validated, but some are not. The attribute 'valid_Landsat' indicates if the observation is validated (TRUE), not validated (FALSE) or could not be validated (NA). Points are in WGS84 (EPSG:4326) - convert to ETRS89-extended / LAEA Europe (EPSG:3035) for better area calculations.


```{r}
points <- read_csv(here("..", "MOTIVATE_validation", "data", "clean", "data_obs_in_val_plots_20260416.csv")) %>%
  st_as_sf(coords = c("Lon_updated", "Lat_updated"), crs = 4326) %>%
  st_transform(3035)
st_crs(points)$Name
```

Create a 500-m buffer around each point.

```{r}
buffers <- points %>% st_buffer(dist = 500, nQuadSegs = 100) # nQuadSegs to make smoother circles
st_crs(buffers)$Name
```

Check that buffer areas are ok (should be around 785398 m² for a 500-m radius circle).

```{r}
buffers$buffer_area <- as.numeric(sf::st_area(buffers))
```

Load raster files with Mahalanobis distance. 

```{r}
raster_files <- list.files(
  here("..", "..", "..", "Data", "MOTIVATE", "MOTIVATE_RS_data", "mapping", "mahalanobis_260508_max_1pixel_expBuffer"),
  pattern = "\\.tif$", full.names = TRUE, recursive = TRUE)
```

# Reproject rasters

Rasters are in WGS84 (EPSG:4326) - convert to ETRS89-extended / LAEA Europe (EPSG:3035) for better area calculations.

```{r message=FALSE}
# ---- OUTPUT FOLDERS ----
output_dir <- here("..", "..", "..", "Data", "MOTIVATE", "MOTIVATE_RS_data", "mapping", "mahalanobis_reproj")

# Create output folder if needed
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

# ---- LOOP OVER RASTERS ----
for (f in raster_files) {
  
  # Build output filename FIRST
  out_name <- file.path(
    output_dir,
    paste0(tools::file_path_sans_ext(basename(f)), "_LAEA.tif")
  )
  
  # ✅ Skip if already exists
  if (file.exists(out_name)) {
    message("Skipping (already exists): ", basename(out_name))
    next
  }
  
  # Otherwise process
  r <- rast(f)
  
  r_laea <- project(r, "EPSG:3035")
  
  writeRaster(r_laea, out_name, overwrite = FALSE)
}
```

# Load reprojected rasters

First remove old rasters in WGS84.

```{r}
rm(raster_files)
```

Now load new reprojected rasters in ETRS89.

```{r}
raster_files_proj <- list.files(
  here("..", "..", "..", "Data", "MOTIVATE", "MOTIVATE_RS_data", "mapping", "mahalanobis_reproj"), 
  pattern = "\\.tif$", full.names = TRUE, recursive = TRUE)
```

```{r}
crs(rast(raster_files_proj[[1]]), describe = TRUE)
```

# Get habitat exents from each raster

## Extract PlotObservationID from raster filenames

```{r}
get_id <- function(x) {
  str_extract(x, "(?<=Obs)\\d+") %>% as.numeric()
}
```

## Define percentiles to use

```{r}
percentiles <- c(0.5, 0.7, 0.9)
```

## Function to process each raster

```{r}
process_raster_fractional <- function(f) {
  
  # -------------------------------------------------------------
  # 1. IDENTIFY PLOT AND LOAD RASTER
  # -------------------------------------------------------------
  # Extract PlotObservationID from filename
  id <- get_id(f)
  
  # Load raster (Mahalanobis distance surface for one plot)
  r <- rast(f)
  
  
  # -------------------------------------------------------------
  # 2. SELECT CORRESPONDING BUFFER
  # -------------------------------------------------------------
  # Each raster corresponds to one PlotObservationID → subset matching buffer
  buf <- buffers %>%
    dplyr::filter(PlotObservationID == id)
  
  # If no matching buffer found, exit function
  if (nrow(buf) == 0) return(NULL)
  
  # Ensure CRS compatibility between raster and buffer
  buf <- st_transform(buf, crs(r))
  st_crs(buf) <- st_crs(r)
  
  
  # -------------------------------------------------------------
  # 3. FRACTIONAL EXTRACTION (FOR STATISTICS)
  # -------------------------------------------------------------
  # Convert terra raster → raster object for exactextractr
  r_raster <- raster::raster(r)
  
  # Extract pixel values within buffer using fractional coverage
  # Returns:
  #  - value: pixel Mahalanobis distance
  #  - coverage_fraction: proportion of pixel inside buffer
  df <- suppressWarnings(
    exact_extract(r_raster, buf)[[1]]
  ) %>%
    dplyr::filter(!is.na(value))
  
  if (nrow(df) == 0) return(NULL)
  
  
  # -------------------------------------------------------------
  # 4. COMPUTE BASIC METRICS
  # -------------------------------------------------------------
  
  # Area of a full raster pixel (m²)
  pixel_area <- prod(res(r))
  
  # Fractional number of pixels in buffer
  total_weight <- sum(df$coverage_fraction)
  
  # True buffer area (based on raster coverage, not vector geometry)
  area_buffer <- total_weight * pixel_area
  
  
  # -------------------------------------------------------------
  # 5. COMPUTE THRESHOLDS
  # -------------------------------------------------------------
  # Thresholds are percentiles of Mahalanobis distance distribution
  thr <- quantile(df$value, probs = percentiles, na.rm = TRUE)
  
  
  # -------------------------------------------------------------
  # 6. PREPARE RASTER (OPTIMIZED)
  # -------------------------------------------------------------
  # ✅ Crop and mask raster to buffer → reduces computation massively
  r_crop <- crop(r, vect(buf))
  buf_vect <- terra::vect(buf)[, 0]  # ✅ drop ALL attributesr
  r_mask <- mask(r_crop, buf_vect)
  
  
  # -------------------------------------------------------------
  # 7. LOOP OVER THRESHOLDS (PERCENTILES)
  # -------------------------------------------------------------
  results <- purrr::map_dfr(seq_along(thr), function(i) {
    
    threshold <- thr[i]          # Mahalanobis threshold value
    perc_val <- percentiles[i]  # percentile (e.g. 0.5, 0.7, 0.9)
    
    
    # -----------------------------------------------------------
    # 7.1 CLASSIFY RASTER INTO HABITAT / NON-HABITAT (FAST)
    # -----------------------------------------------------------
    # NOTE:
    # value ≥ threshold = more similar → habitat
    # value < threshold = less similar → non-habitat
    # ✅ classification done in raster space (much faster)
    r_class <- ifel(r_mask >= threshold, 1L, 0L)
    
    
    # -----------------------------------------------------------
    # 7.2 CONVERT CLASSIFIED RASTER TO POLYGONS
    # -----------------------------------------------------------
    # ✅ Only final classes are polygonized → very few polygons
    pol_sf_thr <- as.polygons(r_class, dissolve = TRUE) %>%
      st_as_sf() %>%
      dplyr::rename(extent = 1) %>%   # rename raster value column
      dplyr::select(extent, geometry) # ✅ KEEP ONLY THESE

    
    # -----------------------------------------------------------
    # 7.3 CLIP POLYGONS TO BUFFER (IMPORTANT)
    # -----------------------------------------------------------
    # ✅ Removes pixel parts outside buffer
    # ✅ Restores true circular boundary
    pol_sf_thr <- suppressWarnings(
      st_intersection(pol_sf_thr, buf)
      )
    
    # ✅ VERY IMPORTANT: remove buffer attributes introduced by intersection
    pol_sf_thr <- pol_sf_thr %>%
      dplyr::select(extent, geometry)
    
    pol_sf_thr <- st_make_valid(pol_sf_thr)
    
    
    # -----------------------------------------------------------
    # 7.4 COMPUTE AREAS (FRACTIONAL, FROM EXACTEXTRACTR)
    # -----------------------------------------------------------
    # IMPORTANT:
    # Use exactextractr-based areas for correct fractional accounting
    
    hab_weight <- sum(df$coverage_fraction[df$value >= threshold], na.rm = TRUE)
    non_weight <- sum(df$coverage_fraction[df$value < threshold], na.rm = TRUE)
    
    area_hab <- hab_weight * pixel_area
    area_non <- non_weight * pixel_area
    
    
    # -----------------------------------------------------------
    # 7.5 RETURN RESULTS
    # -----------------------------------------------------------
    tibble(
      PlotObservationID = id,
      percentile = perc_val,
      threshold_dist = threshold,
      n_pix_buffer = total_weight,
      n_pix_extent = hab_weight,
      n_pix_out = non_weight,
      buffer_m2 = area_buffer,
      extent_m2 = area_hab,
      out_m2 = area_non,
      extent_perc = area_hab / area_buffer * 100
    ) %>%
      dplyr::mutate(geometry = list(pol_sf_thr))
  })
  
  
  # -------------------------------------------------------------
  # 8. RETURN OUTPUT
  # -------------------------------------------------------------
  return(results)
}
```


```{r}
# load existing results
results_all <- readRDS(here("output", "extent_results_fractional.rds"))

# IDs already processed
done_ids <- unique(results_all$PlotObservationID)

# IDs from rasters
raster_ids <- sapply(raster_files_proj, get_id)

# select rasters not yet processed
rasters_to_run <- raster_files_proj[!raster_ids %in% done_ids]

# Print what will run
message("New rasters to process: ", length(rasters_to_run))

# Show skipped rasters
message("Already processed: ", length(done_ids))

# run function only on new ones
new_results <- purrr::map_dfr(
  rasters_to_run,
  process_raster_fractional
)

# combine
results_all <- dplyr::bind_rows(results_all, new_results)

# save updated results
saveRDS(results_all, here("output", "extent_results_fractional.rds"))



# Safety check
if (length(rasters_to_run) == 0) {
  message("Nothing to process!")
}
```

## Get results table

```{r}
results_tbl <- results_all %>%
  dplyr::select(-geometry)
```

```{r}
results_tbl
```

# Export to shapefiles

## Get geomtries

```{r}
# geometries (unnest)
all_polygons <- results_all %>%
  tidyr::unnest(geometry) %>%
  sf::st_as_sf() %>%
  dplyr::select(
    PlotObservationID,
    percentile,
    threshold_dist,
    n_pix_buffer,
    n_pix_extent,
    n_pix_out,
    buffer_m2,
    extent_m2,
    out_m2,
    extent_perc,
    extent,
    geometry
    )
```

## Define directory to save shapefiles

```{r}
out_shapes <- here("output", "shapefiles")
dir.create(out_shapes, recursive = TRUE, showWarnings = FALSE)
```

## Export

```{r}
walk(unique(all_polygons$percentile), function(p) {
  
  shp <- all_polygons %>%
    dplyr::filter(percentile == p)
  
  # Adapt names for shapefile
  names(shp) <- make.names(names(shp), unique = TRUE)
  names(shp) <- substr(names(shp), 1, 10)
  
  out_name <- file.path(
    out_shapes,
    paste0("extent_p", p * 100, ".shp")
  )
  
  st_write(
    shp,
    out_name,
    delete_layer = TRUE,
    quiet = TRUE
  )
})
```

# Maps of extent with different percentiles

Prepare tmap mode:

```{r}
tmap_mode("plot")  # for exporting
```

```{r message=FALSE}
# output folder
output_dir <- "maps_extent"
dir.create(output_dir, showWarnings = FALSE)

ids <- points %>% 
  distinct(PlotObservationID) %>%
  filter(PlotObservationID %in% results_all$PlotObservationID) %>%
  pull(PlotObservationID)

for (id in ids) {
  
  # define output file
  info <- points %>%
    st_drop_geometry() %>%
    dplyr::filter(PlotObservationID == id) %>%
    dplyr::slice(1)
  
  outfile <- file.path(
    output_dir,
    paste0(
      "plot", info$plotID_AV_num,
      "_year", info$year,
      "_obs", info$PlotObservationID,
      ".png"
      )
    )
  
  # skip if already exists
  if (file.exists(outfile)) {
    message("Skipping existing map: ", outfile)
    next
  }
  
  # subset vector layers
  pt  <- points[points$PlotObservationID == id, ]
  buf <- buffers[buffers$PlotObservationID == id, ]
  
  # subset results
  res_id <- results_all %>%
    dplyr::filter(PlotObservationID == id)
  
  # title
  title_txt <- paste(
    paste("Plot", info$plotID_AV_num),
    paste("Year", info$year),
    paste("Obs", info$PlotObservationID),
    paste("EUNIS", info$EUNIS1),
    paste("Valid", ifelse(info$valid_Landsat, "yes", "no")),
    sep = " | "
    )
  
  # ---------------------------------------------------------
  # ✅ 1. BASE MAP (no polygons)
  # ---------------------------------------------------------
  map_base <- tm_shape(buf) +
    tm_tiles("Esri.WorldImagery") +
    tm_borders(col = "red", lwd = 2) +
    tm_shape(pt) +
    tm_symbols(col = "white", fill = "white", size = 0.5) +
    tm_layout(frame = FALSE, legend.show = FALSE) +
    tm_title(title_txt, size = 0.9)
  
  # ---------------------------------------------------------
  # ✅ 2–4. HABITAT MAPS (only extent = 1)
  # ---------------------------------------------------------
  maps_extent <- lapply(sort(unique(res_id$percentile)), function(p) {
    
    pol_sf <- res_id %>%
      dplyr::filter(percentile == p) %>%
      dplyr::pull(geometry) %>%
      .[[1]]
    
    # ✅ keep ONLY habitat
    pol_sf_hab <- pol_sf %>%
      dplyr::filter(extent == 1)
    
    tm_shape(buf) +
      tm_tiles("Esri.WorldImagery") +
      tm_borders(col = "red", lwd = 2) +
      
      # ✅ habitat only
      tm_shape(pol_sf_hab) +
      tm_fill(
        col = "gold1",   
        alpha = 0.6
      ) +
      
      tm_shape(pt) +
      tm_symbols(col = "white", fill = "white", size = 0.5) +
      
      tm_layout(
        frame = FALSE,
        legend.show = FALSE   # ✅ no legend
      ) +
      
      tm_title(paste0("Percentile: ", p), size = 0.9)
  })
  
    # ---------------------------------------------------------
  # ✅ Combine into 2x2 grid
  # ---------------------------------------------------------
  final_map <- tmap_arrange(
    map_base, maps_extent[[1]], maps_extent[[2]], maps_extent[[3]],
    ncol = 2, nrow = 2)
  
  # save figure
  
  tmap_save(final_map, outfile, width = 8, height = 8, units = "in", dpi = 300)
}
```

# Session info

```{r}
sessionInfo()
```

